---
title: "From DiD to SDID: A Ladder of Synthetic Control Estimators"
subtitle: "Seven ways to build a counterfactual United Kingdom, hand-coded and then packaged"
author: "Carlos Mendez"
date: "2026-07-31"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: darkly
    fig-width: 9
    fig-height: 6
    fig-dpi: 300
execute:
  warning: false
  message: false
  freeze: auto
---

## What this notebook does

The United Kingdom voted to leave the European Union on 23 June 2016. There is only
one United Kingdom, so the cost of that decision can only be measured against a
country that never existed. This notebook builds that country seven times over,
climbing the ladder of single-treated-unit estimators:

| Rung | Estimator | Unit weights | Time weights | Unit effect |
|---|---|---|---|---|
| 0 | Difference-in-differences | fixed at $1/J$ | fixed | yes |
| 1 | Synthetic control | optimised on the simplex | none | **no** |
| 2 | Demeaned SC | optimised on demeaned data | fixed | yes |
| 3 | Synthetic DiD | same as DSC | **optimised** | yes |
| 4 | MASC | cross-validated blend with matching | none | no |
| 5 | Augmented SC | SC weights + ridge correction | none | no |

Every rung is solved from scratch first, then reproduced with its package.

The companion post, with the full narrative and all eighteen figures, is at
<https://carlos-mendez.org/post/r_sc_dsc_sdid/>. The canonical script is
`analysis.R` in this folder.

## Setup

```{r}
#| label: setup-packages
required <- c("quadprog", "ggplot2", "patchwork")
missing  <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")

suppressPackageStartupMessages({
  library(quadprog); library(ggplot2); library(patchwork)
})

has <- function(p) requireNamespace(p, quietly = TRUE)

# The three GitHub-only packages. Skipped gracefully if absent -- every estimate
# below is also computed by hand, so the notebook renders either way.
#   remotes::install_github("synth-inference/synthdid")
#   remotes::install_github("ebenmichael/augsynth")
#   masc needs its Gurobi dependency stripped first; see analysis.R.

set.seed(20260801)

STEEL <- "#6a9bcc"; ORANGE <- "#d97757"; TEAL <- "#00d4c8"
BG <- "#0f1729"; PANEL <- "#1f2b5e"; TXT <- "#c8d0e0"; TXT2 <- "#e8ecf2"
GREY <- "#54618a"; MUTED <- "#8b9dc3"; GOLD <- "#e8b04b"

theme_site <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(plot.background  = element_rect(fill = BG, colour = NA),
          panel.background = element_rect(fill = BG, colour = NA),
          panel.grid.major = element_line(colour = PANEL, linewidth = 0.3),
          panel.grid.minor = element_blank(),
          text = element_text(colour = TXT), axis.text = element_text(colour = TXT),
          plot.title = element_text(colour = TXT2, face = "bold"),
          plot.subtitle = element_text(colour = TXT),
          plot.caption = element_text(colour = MUTED, size = base_size - 3, hjust = 0),
          legend.text = element_text(colour = TXT),
          legend.title = element_text(colour = TXT2),
          legend.key = element_rect(fill = BG, colour = NA),
          strip.text = element_text(colour = TXT2, face = "bold"),
          strip.background = element_rect(fill = PANEL, colour = NA))
}
theme_set(theme_site())
```

## The data

Quarterly log real GDP for 24 OECD economies, 1995Q1 to 2020Q4, indexed so each
country's 1995 average equals one. The United Kingdom is the treated unit and the
other 23 are the donor pool.

```{r}
#| label: data-download
url <- paste0("https://raw.githubusercontent.com/cmg777/starter-academic-v501/",
              "master/content/post/r_sc_dsc_sdid/brexit_analysis.csv")
if (!file.exists("brexit_analysis.csv")) download.file(url, "brexit_analysis.csv")
panel <- read.csv("brexit_analysis.csv", stringsAsFactors = FALSE)

CT     <- unique(panel$country[order(panel$unit_id)])
Y      <- matrix(panel$log_rgdp[order(panel$unit_id, panel$t)],
                 nrow = 104, dimnames = list(NULL, CT))
QLAB   <- panel$quarter_label[panel$unit_id == 1][order(panel$t[panel$unit_id == 1])]
QDEC   <- panel$date_dec[panel$unit_id == 1][order(panel$t[panel$unit_id == 1])]
UK     <- which(CT == "United Kingdom")
DONORS <- setdiff(seq_along(CT), UK)

# CAREFUL: T0 here is the NUMBER of pre-treatment quarters, not a period index.
# The treatment materialises in 2016Q3, which is t = 87.
T0 <- 86
E  <- c(`2018Q4` = 96, `2019Q4` = 100)
Z1 <- Y[1:T0, UK]; Z0 <- Y[1:T0, DONORS]

cat(sprintf("%d countries x %d quarters (%s to %s); %d donors; %d pre-treatment quarters\n",
            ncol(Y), nrow(Y), QLAB[1], QLAB[104], length(DONORS), T0))
```

## The UK inside the donor cloud

```{r}
#| label: fig-gdp-paths
long <- data.frame(date = rep(QDEC, 24), country = rep(CT, each = 104),
                   y = as.vector(Y))
long$uk <- long$country == "United Kingdom"
ggplot(subset(long, !uk), aes(date, y, group = country)) +
  geom_line(colour = GREY, linewidth = 0.35, alpha = 0.75) +
  geom_line(data = subset(long, uk), colour = ORANGE, linewidth = 1.1) +
  geom_vline(xintercept = 2016.50, linetype = "dashed", colour = TEAL) +
  labs(title = "The United Kingdom is one line in a crowd",
       subtitle = "Log real GDP (1995 average = 1); orange is the UK",
       x = NULL, y = "log real GDP index")
```

No single donor tracks the UK, which is why we build a blend. But the 2008-09
collapse is synchronised across every country, and that common movement is what
the demeaned and time-weighted estimators exploit.

## One solver, five jobs

Every rung reduces to the same problem: minimise $\lVert b - Aw \rVert^2$ subject
to $w \ge 0$ and $\sum_k w_k = 1$. Write it once.

```{r}
#| label: fit-solver
simplex_ls <- function(A, b, ridge = 1e-10) {
  k <- ncol(A)
  w <- solve.QP(crossprod(A) + ridge * diag(k), crossprod(A, b),
                cbind(rep(1, k), diag(k)), c(1, rep(0, k)), meq = 1)$solution
  w[w < 1e-10] <- 0
  w / sum(w)
}

loss <- function(w, b = 0) -100 * (Y[E, UK] - Y[E, DONORS] %*% w - b)
res  <- data.frame()
add  <- function(m, w, b = 0, note = "") {
  v <- loss(w, b)
  res <<- rbind(res, data.frame(method = m, `2018Q4` = round(v[1], 3),
                                `2019Q4` = round(v[2], 3), note = note,
                                check.names = FALSE))
  invisible(NULL)
}
```

## Rung 0 — Difference-in-differences

Every donor counts the same; a unit fixed effect absorbs the level gap.

```{r}
#| label: fit-did
w_did <- rep(1 / length(DONORS), length(DONORS))
b_did <- mean(Z1 - Z0 %*% w_did)
add("DiD", w_did, b_did, "uniform weights")
cat(sprintf("pre-treatment RMSE: %.5f\n", sqrt(mean((Z1 - Z0 %*% w_did - b_did)^2))))
```

Nearly 5% — roughly double every other rung, and its pre-treatment fit error is
four times synthetic control's. DiD is fitting a trend divergence that began years
before the referendum.

## Rung 1 — Synthetic control

$$\hat{\omega}^{sc} = \arg\min_{\omega \in \mathbb{W}} \sum_{t=1}^{T_0} \Big( y_{1,t} - \sum_j \omega_j y_{j,t} \Big)^2$$

```{r}
#| label: fit-sc
w_sc <- simplex_ls(Z0, Z1)
add("SC (hand-coded)", w_sc, 0, "exact quadratic program")

top <- data.frame(country = CT[DONORS], omega = round(w_sc, 4))
print(head(top[order(-top$omega), ], 8), row.names = FALSE)
cat(sprintf("\npre-treatment RMSPE: %.5f\n", sqrt(mean((Z1 - Z0 %*% w_sc)^2))))
```

```{r}
#| label: fig-sc-fit
syn <- as.vector(Y[, DONORS] %*% w_sc)
df  <- data.frame(date = QDEC, uk = Y[, UK], syn = syn, gap = Y[, UK] - syn)

pA <- ggplot(df, aes(date)) +
  geom_line(aes(y = uk), colour = ORANGE, linewidth = 1) +
  geom_line(aes(y = syn), colour = STEEL, linewidth = 1, linetype = "22") +
  geom_vline(xintercept = 2016.50, linetype = "dashed", colour = TEAL) +
  labs(title = "The UK and its synthetic control", x = NULL, y = "log real GDP index")
pB <- ggplot(df, aes(date, gap)) +
  geom_hline(yintercept = 0, colour = MUTED) +
  geom_vline(xintercept = 2016.50, linetype = "dashed", colour = TEAL) +
  geom_line(colour = STEEL, linewidth = 0.9) +
  labs(x = NULL, y = "gap (log points)")
pA / pB + plot_layout(heights = c(2, 1.2))
```

## Rung 2 — Demeaned synthetic control

SC has no intercept, so it rejects a blend that moves in perfect parallel but sits
at a different level. Demean first, add the offset back as a constant.

```{r}
#| label: fit-dsc
w_dsc <- simplex_ls(sweep(Z0, 2, colMeans(Z0)), Z1 - mean(Z1))
b_dsc <- mean(Z1 - Z0 %*% w_dsc)
add("DSC (hand-coded)", w_dsc, b_dsc, sprintf("offset %+.5f", b_dsc))
cat(sprintf("correlation of SC and DSC weights: %.4f\n", cor(w_sc, w_dsc)))
```

The offset is tiny, which is the finding: the synthetic control was already
level-balanced, so SC was not sacrificing shape to chase level.

## Rung 3 — Synthetic difference-in-differences

The time-weight problem is the unit-weight problem run on the transpose.

- $\omega$ asks which **countries**, blended, reproduce the UK.
- $\lambda$ asks which **quarters**, blended, reproduce the treatment quarter.

Note the two different demeanings: DSC removes each *country's* time-series mean;
the $\lambda$ problem removes each *quarter's* cross-sectional mean.

```{r}
#| label: fit-sdid
# A has donors as ROWS and quarters as COLUMNS, each quarter demeaned across donors.
A_lam  <- sweep(t(Z0), 2, colMeans(t(Z0)))
lambda <- simplex_ls(A_lam, Y[T0 + 1, DONORS] - mean(Y[T0 + 1, DONORS]))
b_sdid <- drop(lambda %*% (Z1 - Z0 %*% w_dsc))
add("SDID (hand-coded)", w_dsc, b_sdid,
    sprintf("%.0f%% of lambda on %s", 100 * lambda[T0], QLAB[T0]))

cat("Top time weights:\n")
top_l <- order(-lambda)[1:4]
for (k in top_l) if (lambda[k] > 1e-4) cat(sprintf("  %-8s %.4f\n", QLAB[k], lambda[k]))
cat(sprintf("\nDSC offset %+.5f  vs  SDID offset %+.5f\n", b_dsc, b_sdid))
```

```{r}
#| label: fig-lambda
ld <- data.frame(date = QDEC[1:T0], lambda = lambda)
rw <- data.frame(prev = as.vector(Z0[1:(T0-1), ]), cur = as.vector(Z0[2:T0, ]))

pL <- ggplot(ld, aes(date, lambda)) +
  geom_hline(yintercept = 1/T0, colour = GOLD, linetype = "22") +
  geom_segment(aes(xend = date, yend = 0), colour = STEEL) +
  geom_point(colour = STEEL, size = 1.5) +
  labs(title = "Time weights collapse onto the last pre-treatment quarter",
       subtitle = "Gold line: the uniform (two-way fixed effects) alternative",
       x = NULL, y = "lambda")
pR <- ggplot(rw, aes(prev, cur)) +
  geom_abline(colour = ORANGE, linetype = "22") +
  geom_point(colour = STEEL, alpha = 0.3, size = 0.8) +
  labs(title = "Why: log GDP is nearly a random walk",
       subtitle = sprintf("correlation %.5f", cor(rw$prev, rw$cur)),
       x = "quarter t-1", y = "quarter t")
pL / pR
```

If the best predictor of next quarter is this quarter, then the best blend of past
quarters is the most recent one alone. The collapse is a property of GDP, not a bug.

## Rung 4 — MASC

A cross-validated blend of $m$-nearest-neighbour matching and synthetic control.

```{r}
#| label: fit-masc
nn <- function(Z0, Z1, m) {
  d <- colSums((Z1 - Z0)^2)
  s <- d %in% sort(d)[1:m]
  as.numeric(s) / sum(s)
}

# Rolling origin. NOTE: folds must run from 6 to T0-1. Passing masc's
# `min_preperiods` to current package master yields only 5 folds, selects
# phi = 1 (pure matching) and does not reproduce the published result.
set_f <- 6:(T0 - 1)
wt    <- rep(1 / length(set_f), length(set_f))
ysc   <- vapply(set_f, function(k) drop(Z0[k+1, ] %*% simplex_ls(Z0[1:k, ], Z1[1:k])), 0)
ytr   <- Z1[set_f + 1]
cv <- do.call(rbind, lapply(1:10, function(m) {
  ymt <- vapply(set_f, function(k) drop(Z0[k+1, ] %*% nn(Z0[1:k, ], Z1[1:k], m)), 0)
  phi <- min(1, max(0, drop((wt*(ytr-ysc)) %*% (ymt-ysc) / ((wt*(ymt-ysc)) %*% (ymt-ysc)))))
  data.frame(m = m, phi = phi, cv = sum(wt * (ytr - phi*ymt - (1-phi)*ysc)^2))
}))
best   <- cv[which.min(cv$cv), ]
w_masc <- best$phi * nn(Z0, Z1, best$m) + (1 - best$phi) * w_sc
add("MASC (hand-coded)", w_masc, 0, sprintf("m = %d, phi = %.3f", best$m, best$phi))
print(transform(cv, phi = round(phi, 4), cv = signif(cv, 4)), row.names = FALSE)
```

```{r}
#| label: fig-masc-cv
ggplot(cv, aes(factor(m), cv)) +
  geom_col(fill = STEEL, width = 0.7) +
  geom_col(data = best, fill = TEAL, width = 0.7) +
  geom_text(aes(label = sprintf("phi=%.2f", phi)), colour = TXT, vjust = -0.5, size = 3) +
  labs(title = "MASC cross-validates the blend rather than guessing it",
       x = "m (matched donors)", y = "rolling-origin CV error")
```

## Rung 5 — Augmented synthetic control

Drop the non-negativity constraint; a ridge penalty pulls the weights back toward
the SC solution. Weights may go negative.

The ridge parameter is not ours to invent — `augsynth` picks it by
leave-one-period-out cross-validation. Feed its choice back into the closed form
and the two agree exactly.

```{r}
#| label: fit-ascm
ascm <- function(lr) {
  Xc  <- sweep(t(Z0), 2, colMeans(t(Z0)))
  x1c <- Z1 - colMeans(t(Z0))
  as.numeric(w_sc + solve(tcrossprod(Xc) + lr * diag(length(DONORS)),
                          Xc %*% (x1c - crossprod(Xc, w_sc))))
}
w_as <- ascm(0.13858)          # the value augsynth's CV selects on this panel
add("ASCM (hand-coded)", w_as, 0, sprintf("%d negative weights", sum(w_as < 0)))
neg <- data.frame(country = CT[DONORS], omega = round(w_as, 4))
print(head(neg[order(neg$omega), ], 5), row.names = FALSE)
```

## The same rungs, via their packages

```{r}
#| label: fit-packages
if (has("synthdid")) {
  Ysd  <- t(cbind(Y[1:(T0+1), DONORS], Y[1:(T0+1), UK]))   # units x time, treated LAST
  p_sc <- synthdid::synthdid_estimate(Ysd, N0 = 23, T0 = T0, zeta.omega = 0,
                                      zeta.lambda = 0, omega.intercept = FALSE,
                                      lambda.intercept = FALSE)
  p_sd <- synthdid::synthdid_estimate(Ysd, N0 = 23, T0 = T0, zeta.omega = 0,
                                      zeta.lambda = 0, omega.intercept = TRUE,
                                      lambda.intercept = TRUE)
  o_sc <- attr(p_sc, "weights")$omega
  o_sd <- attr(p_sd, "weights")$omega
  l_sd <- attr(p_sd, "weights")$lambda
  add("SC (synthdid)",   o_sc)
  add("DSC (synthdid)",  o_sd, mean(Z1 - Z0 %*% o_sd))
  add("SDID (synthdid)", o_sd, drop(l_sd %*% (Z1 - Z0 %*% o_sd)))
  cat(sprintf("hand-coded vs package, SC weights: max abs difference %.3e\n",
              max(abs(w_sc - o_sc))))
}

if (has("augsynth")) {
  ad <- data.frame(unitnum = rep(1:24, each = T0 + 1), t = rep(1:(T0 + 1), 24),
                   value = as.vector(Y[1:(T0 + 1), c(UK, DONORS)]))
  ad$treatment <- as.integer(ad$unitnum == 1 & ad$t == T0 + 1)
  a <- augsynth::augsynth(value ~ treatment, unitnum, t, ad,
                          progfunc = "Ridge", scm = TRUE)
  add("ASCM (augsynth)", as.numeric(a$weights), 0,
      sprintf("CV lambda %.5f", a$lambda))
  cat(sprintf("hand-coded vs package, ASCM weights: max abs difference %.3e\n",
              max(abs(ascm(a$lambda) - as.numeric(a$weights)))))
}
```

The hand-coded SC and the package differ by about 0.016 on individual weights and
0.02 percentage points on the estimate. That is not a bug: with no ridge penalty
the objective is nearly flat (condition number around $7 \times 10^5$), so
`synthdid`'s Frank-Wolfe optimiser stops on its iteration cap rather than at the
optimum. Section 8.4 of the post works through the whole convergence ladder.

## The ladder, side by side

```{r}
#| label: results-table
print(res, row.names = FALSE)
cat("\nBorn et al. (2019) reported 2.4% at 2018Q4 and 3.6% at 2019Q4.\n")
```

```{r}
#| label: fig-ladder
main <- res[!grepl("hand-coded", res$method) | !grepl("SC \\(|DSC \\(|SDID \\(|ASCM \\(", res$method), ]
main$method <- factor(main$method, levels = rev(main$method))
ggplot(main, aes(`2018Q4`, method)) +
  geom_vline(xintercept = 2.4, colour = ORANGE, linetype = "22") +
  geom_point(colour = STEEL, size = 3.2) +
  geom_text(aes(label = sprintf("%.2f", `2018Q4`)), vjust = -1, colour = TXT, size = 3) +
  labs(title = "Every rung puts the Brexit cost above 2.4%",
       subtitle = "Estimated 2018Q4 GDP shortfall, treatment dated 2016Q3, no covariates",
       x = "GDP loss (%)", y = NULL,
       caption = "Dashed line: Born et al. (2019).")
```

## Where to go next

The post covers four things this notebook leaves out: the extrapolation-versus-
interpolation bias decomposition that motivates the whole ladder, the in-sample
placebo tournament over twenty artificial treatment dates, the covariates verdict,
and permutation inference. Run `analysis.R` in this folder to reproduce all
eighteen figures and every table.

<https://carlos-mendez.org/post/r_sc_dsc_sdid/>
