---
title: "Code for Replication in R and Stata"
format: 
  html:
    page-layout: article
    code-fold: show
    code-tools: true
    toc: true
    toc-location: left
    toc-depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    toc-expand: 3
    self-contained: true
    theme: zephyr
    grid:
      margin-width: 100px
      body-width: 1200px
---

# Introduction

In this appendix we provide the code to replicate the tables and figures in the article. By pressing the $\triangleright$ code button you can see the R code that creates the relevant figure or table, with a description of the steps involved in computing the analyses.

# Table 1: Medicaid Expansion Adoption

The following code block produces **Table 1**, which shows the roll-out of Medicaid Expansion under the Affordable Care Act (ACA) across states.

```{r, include = FALSE}
# we need the package that integrates stata before we can run it
library(Statamarkdown)
library(readxl)
library(knitr)
library(readr)
library(purrr)
knitr::opts_chunk$set(engine.opts = list(cleanup = TRUE))

```

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE, results = 'asis'}
## Install packages from Github if needed
# load packages
library(tidyverse)
library(kableExtra)
library(modelsummary)
library(gt)
library(did)
library(ggthemes)
library(fixest)
library(latex2exp)
library(HonestDiD)

# omit NA from tables
options(knitr.kable.NA = '')

# set baseline output for modelsummary
options(modelsummary_factory_default = 'kableExtra')

# set ggplot theme
theme_set(
  theme_clean() + 
    theme(plot.background = element_blank(),
          legend.background = element_rect(color = "white"),
          strip.background = element_rect(color = "black"))
)

# load cleaned data
mydata <- read_csv(here::here("data", "county_mortality_data.csv")) %>% 
  # make state the abbreviation
  mutate(state = str_sub(county, nchar(county) - 1, nchar(county))) %>%
  # drop District of Columbia from the data
  filter(state != "DC") %>% 
  # make a second adoption variable for the table
  mutate(adopt = case_when(
    # missing is non-expansion
    is.na(yaca) ~ "Non-Expansion",
    # fix a couple pre-2014 adoptions from Miller, Johnson and Wherry
    state %in% c("DE", "MA", "NY", "VT") ~ "Pre-2014", 
    TRUE ~ as.character(yaca)
  ))
  
# get adoption year by state
adopts <- mydata %>% 
  select(state, adopt) %>% 
  distinct() %>% 
  arrange(state)

# first get the share of states, share of counties, and share of adults in 2013 by adoption category
# first the states and share of states
states <- adopts %>% 
  group_by(adopt) %>% 
  summarize(states = paste0(state, collapse = ", "),
            state_share = length(state) / 50)

# next get the county share and the population share
counties_pop <- mydata %>% 
  # just for year 2013
  filter(year == 2013) %>% 
  # get county and population totals
  mutate(total_counties = n(),
         total_pop = sum(population_20_64, na.rm = TRUE)) %>% 
  group_by(adopt) %>% 
  # make into shares
  summarize(county_share = n() / mean(total_counties),
            pop_share = sum(population_20_64, na.rm = TRUE) / mean(total_pop))
  
# make a table and export
states %>% 
  left_join(counties_pop, by = "adopt") %>% 
  slice(9, 1:8) %>% 
  # format the numbers to two digits
  mutate(across(state_share:pop_share, ~ scales::number(., accuracy = 0.01))) %>%
  # format the table
  kable(
    col.names = c("Expansion \n Year", "States", "Share of States", "Share of Counties", 
                  "Share of Adults (2013)"),
    booktabs = T, caption = "Medicaid Expansion Under the Affordable Care Act",
    label = "adoptions",
    align = c("c"),
    escape = FALSE,
    linesep = ""
  ) %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position")) %>% 
  column_spec(2, width = "20em")
```

### Stata

```{stata, eval = FALSE}
clear
capture log close

******************************************************************************
* 1. Load original data 
******************************************************************************
	cd "../../data"
  insheet using "county_mortality_data.csv", clear

	keep state county_code county year yaca population_20_64
	
	*Define treatment timing
	gen treat_year = real(yaca)
	replace treat_year = 0 if missing(treat_year)
	tostring treat_year, gen(treat_str)
	destring yaca, replace force

	destring population_20_64, replace force
	
	* Drop DC as in JEL Table 1
	drop if state == "District of Columbia"

****************************************************
* 2. Create adoption category                      *
****************************************************
	gen adopt = ""
	replace adopt = "Pre-2014" if inlist(state, "Delaware", "Massachusetts", "New York", "Vermont")
	replace adopt = string(yaca) if !missing(yaca) & adopt == ""
	replace adopt = "Non-Expansion" if missing(yaca)

****************************************************
* 3. Get number of states and state share          *
****************************************************
preserve
gen st_abr = substr(county, max(1, strlen(county)-1), 2)
keep st_abr adopt
duplicates drop
sort adopt st_abr

by adopt: gen strL state_list = st_abr
by adopt: replace state_list = state_list[_n-1] + ", " + st_abr if adopt[_n-1] == adopt[_n]

bysort adopt: gen state_count=_N

gen state_share = state_count / 50

gen len= strlen(state_list)
bysort adopt: egen max=max(len)
keep if len==max 

keep adopt state_list state_share
tempfile state_summary
save `state_summary', replace
restore
****************************************************
* 4. Get county & adult pop share (year = 2013)    *
****************************************************
	preserve
	keep if year == 2013

	* Total county count and total 20–64 pop in 2013
	su population_20_64
	scalar total_pop = r(sum)

	su county_code
	scalar total_counties = r(N)

	collapse (count) county_count=county_code (sum) pop=population_20_64, by(adopt)

	gen county_share = county_count / total_counties
	gen pop_share = pop / total_pop

	tempfile county_summary
	save `county_summary'
	restore

****************************************************
* 5. Merge and order categories for display        *
****************************************************
	use `state_summary', clear
	merge 1:1 adopt using `county_summary', nogen

	* Order like JEL Table 1
	gen order = .
	replace order = 1 if adopt == "Pre-2014"
	replace order = 2 if adopt == "2014"
	replace order = 3 if adopt == "2015"
	replace order = 4 if adopt == "2016"
	replace order = 5 if adopt == "2019"
	replace order = 6 if adopt == "2020"
	replace order = 7 if adopt == "2021"
	replace order = 8 if adopt == "2023"
	replace order = 9 if adopt == "Non-Expansion"
	sort order

	* Round to match JEL formatting
	gen state_share_fmt = string(state_share, "%4.2f")
	gen county_share_fmt = string(county_share, "%4.2f")
	gen pop_share_fmt = string(pop_share, "%4.2f")

****************************************************
* 6. Export table                                 *
****************************************************

export excel adopt state_list state_share_fmt county_share_fmt pop_share_fmt ///
    using "../data/adoptions_table.xlsx", firstrow(variables) replace
restore 
```
:::

# $2\times2$ DiD setups

The following tables and figures show our analyses within the $2\times2$ section of the paper.

## Table 2: Simple $2\times2$ DiD

**Table 2** shows how we can calculate the simple $2\times2$ DiD comparing the change in the crude mortality rates between states that chose to expand Medicaid in 2014 and those that had not yet expanded by 2019.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
# set seed for reproducibility
set.seed(20240924)

# set filepath locations for tables and figures
file.loc.tab <- here::here("tables/")
file.loc.fig <- here::here("figures/")

# load cleaned data
mydata <- read_csv(here::here("data", "county_mortality_data.csv")) %>% 
  # make state the abbreviation
  mutate(state = str_sub(county, nchar(county) - 1, nchar(county))) %>%
  # drop DC and pre-2014 adoption states
  filter(!(state %in% c("DC", "DE", "MA", "NY", "VT"))) %>% 
  # drop states that adopt between 2014 and 2019
  filter(yaca == 2014 | is.na(yaca) | yaca > 2019)

# set the covariates that we're going to use.
covs <- c("perc_female","perc_white", "perc_hispanic", "unemp_rate", "poverty_rate", "median_income")

## Clean data and add in covariates
mydata <- mydata %>% 
  # make variables
  mutate(perc_white = population_20_64_white / population_20_64 * 100,
         perc_hispanic = population_20_64_hispanic / population_20_64 * 100,
         perc_female = population_20_64_female/ population_20_64 * 100,
         unemp_rate = unemp_rate * 100,
         median_income = median_income / 1000) %>% 
  # keep just subset of variables that we will use later
  select(state, county, county_code, year, population_20_64, yaca,
         starts_with("perc_"), crude_rate_20_64, all_of(covs))

# keep only counties with full observations for outcome and covariates in 2013 and 2014
mydata <- mydata %>%
  # allow the aca expansion variable to be missing
  drop_na(!yaca) %>%
  group_by(county_code) %>% 
  # need full covariates for 2013 and 2014
  filter(length(which(year == 2013 | year == 2014)) == 2) %>% 
  ungroup()

# finally, keep only counties with full mortality data for 2009 to 2019
mydata <- mydata %>% 
  group_by(county_code) %>% 
  drop_na(crude_rate_20_64) %>% 
  filter(n() == 11)

# make a smaller dataset with only years 2013 and 2014
short_data <- mydata %>% 
  # make a binary variable that identifies ACA expansion and post-years
  mutate(Treat = if_else(yaca == 2014 & !is.na(yaca), 1, 0),
         Post = if_else(year == 2014, 1, 0)) %>% 
  # filter years for just 2013 and 2014
  filter(year %in% c(2013, 2014)) %>% 
  # make a variable with population weight in 2013
  group_by(county_code) %>% 
  mutate(set_wt = population_20_64[which(year == 2013)]) %>% 
  ungroup()
  
# get group means for Table 2
# unweighted expansion pre-reform
T_pre <- short_data %>% 
  filter(Treat == 1 & year == 2013) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# unweighted non-expansion pre-form 
C_pre <- short_data %>% 
  filter(Treat == 0 & year == 2013) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# unweighted expansion post-reform
T_Post <- short_data %>% 
  filter(Treat == 1 & year == 2014) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# unweighted non-expansion post-reform
C_Post <- short_data %>% 
  filter(Treat == 0 & year == 2014) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# get the same group means but this time weighted
T_pre_weight <- short_data %>% 
  filter(Treat == 1 & year == 2013) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

C_pre_weight <- short_data %>% 
  filter(Treat == 0 & year == 2013) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

T_Post_weight <- short_data %>% 
  filter(Treat == 1 & year == 2014) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

C_Post_weight <- short_data %>% 
  filter(Treat == 0 & year == 2014) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

# function to do a reasonable rounding
g_round <- function(x, k) {format(round(x, k), nsmall = k)}

# make table and export
tribble(
  ~"", ~"Expansion", ~ "No Expansion", ~"Gap/DiD", ~"Expansion", ~"No Expansion", ~"Gap/DiD",
  "2013", as.character(g_round(T_pre, 1)), as.character(g_round(C_pre, 1)), as.character(g_round(T_pre - C_pre, 1)), 
  as.character(g_round(T_pre_weight, 1)), as.character(g_round(C_pre_weight, 1)), as.character(g_round(T_pre_weight - C_pre_weight, 1)),
  "2014", as.character(g_round(T_Post, 1)), as.character(g_round(C_Post, 1)), as.character(g_round(T_Post - C_Post, 1)), 
  as.character(g_round(T_Post_weight, 1)), as.character(g_round(C_Post_weight, 1)), as.character(g_round(T_Post_weight - C_Post_weight, 1)),
  "Trend/DiD", as.character(g_round(T_Post - T_pre, 1)), as.character(g_round(C_Post-C_pre, 1)), 
  as.character(g_round((T_Post - T_pre) - (C_Post-C_pre), 1)),
    as.character(g_round(T_Post_weight - T_pre_weight, 1)), as.character(g_round(C_Post_weight-C_pre_weight, 1)), 
  as.character(g_round((T_Post_weight - T_pre_weight) - (C_Post_weight-C_pre_weight), 1))
) %>% 
  # format table
  kable(align = 'c',
        booktabs = T, 
        escape = F,
        caption = "Simple 2 X 2 DiD") %>% 
  kable_styling() %>% 
  row_spec(3, color = "BrickRed") %>% 
  column_spec(c(1:3, 5:6), color = "black") %>% 
  row_spec(3, italic = TRUE) %>% 
  column_spec(c(4, 7), italic = TRUE) %>% 
  add_header_above(c(" " = 1, "Unweighted Averages" = 3, "Weighted Averages" = 3))
```

### Stata

```{stata, eval = FALSE}

********************************
* 1. Load and prepare data
********************************
	clear all
	set more off
	capture log close

	insheet using "county_mortality_data.csv", clear

********************************
* 2. Prepare timing groups
********************************
	*Drop early and mid adopters
	drop if inlist(state, "District of Columbia", "Delaware", "Massachusetts", "New York", "Vermont")

	*Define treatment timing
	gen treat_year = real(yaca)
	replace treat_year = 0 if missing(treat_year)
	tostring treat_year, gen(treat_str)
	destring yaca, replace force

********************************
* 3. Prepare covariates
********************************
	*destring covariates
	destring deaths population_20_64 crude_rate_20_64 population_total population_20_64_hispanic population_20_64_female population_20_64_white unemployed labor_force unemp_rate poverty_rate median_income, replace force
	ren population_total total_population

	*Drop counties with missing covariates
	gen perc_white    = population_20_64_white   / population_20_64 * 100
	gen perc_hispanic = population_20_64_hispanic/ population_20_64 * 100
	gen perc_female   = population_20_64_female  / population_20_64 * 100
	gen unemp_rate_pc = unemp_rate * 100
	gen median_income_k = median_income/1000
	drop if missing(crude_rate_20_64, population_20_64, ///
					perc_white, perc_hispanic, perc_female, ///
					unemp_rate_pc, poverty_rate, median_income_k)

********************************
* 4. Sample construction
********************************				
	*Keep counties with full 11-year panel
	bys county_code (year): gen panel_n = _N
	bys county_code: keep if panel_n == 11
	drop panel_n 
	replace treat_year=0 if treat_year>2019

********************************
* 5. Construct population weight
********************************
	bys county_code: egen set_wt = max(cond(year==2013, population_20_64, .))

********************************
* 6. Label variables
********************************
	label variable crude_rate_20_64 "Crude Mortality Rate"

	
	
save "did_jel_aca_replication_data", replace

***************************************************************
*1. Table 2: weighted and unweighted averages for the simple 2x2 DiD
***************************************************************
	*Read in data, keep 2013–2014 and 2014 expansion vs "non expansion" states
	use "did_jel_aca_replication_data", clear
	keep if inlist(year, 2013, 2014) & inlist(yaca,2014,2020,2021,2023,.)
	gen Treat = (yaca==2014)		// 1 = ACA in 2014
	gen Post = (year == 2014)     	// 1 = post-expansion year

	*Means
	*unweighted
	sum crude_rate if Treat==1 & year==2014
	global u12014 = r(mean)
	sum crude_rate if Treat==1 & year==2013
	global u12013 = r(mean)
	sum crude_rate if Treat==0 & year==2014
	global u02014 = r(mean)
	sum crude_rate if Treat==0 & year==2013
	global u02013 = r(mean)

	*weighted
	sum crude_rate if Treat==1 & year==2014 [aw=set_wt]
	global w12014 = r(mean)
	sum crude_rate if Treat==1 & year==2013  [aw=set_wt]
	global w12013 = r(mean)
	sum crude_rate if Treat==0 & year==2014  [aw=set_wt]
	global w02014 = r(mean)
	sum crude_rate if Treat==0 & year==2013 [aw=set_wt]
	global w02013 = r(mean)
	


	*Gaps, trends, DiDs 
	* Unweighted
	global Gap13_u = ($u12013 - $u02013)
	global Gap14_u = ($u12014 - $u02014)
	global DiD_u   = ( ($u12014 - $u12013) - ($u02014 - $u02013) )
	global Trend_T_u = ($u12014 - $u12013)
	global Trend_C_u = ($u02014 - $u02013)

	* Weighted
	global Gap13_w = ($w12013 - $w02013)
	global Gap14_w = ($w12014 - $w02014)
	global DiD_w   = ( $w12014 - $w12013 ) - ( $w02014 - $w02013 )
	global Trend_T_w = ($w12014 - $w12013)
	global Trend_C_w = ($w02014 - $w02013)



	*Rounding for the table
	foreach nm in u12013 u02013 u12014 u02014 ///
				 w12013 w02013 w12014 w02014 ///
				 Gap13_u Gap14_u DiD_u Trend_T_u Trend_C_u ///
				 Gap13_w Gap14_w DiD_w Trend_T_w Trend_C_w {
		local `nm' : display %6.1f ${`nm'}
		display("`nm'")
	}

	display("`u12013'")
	display "$u12013"
	* Brick-red DiD cells
	local DiD_u_cell "\em{\em{\textcolor{BrickRed}{`DiD_u'}}}"
	local DiD_w_cell "\em{\em{\textcolor{BrickRed}{`DiD_w'}}}"

	
	
	
/*--------------------------------------------------------------------
  8. Export Excel Table for Rendering
--------------------------------------------------------------------*/
clear
set obs 3

// row labels
gen str10 row = ""
replace row = "2013"      in 1
replace row = "2014"      in 2
replace row = "Trend/DiD" in 3

// columns (numeric); convert locals like `u12013' to numbers with real()
gen double Expansion_unw   = .
gen double NoExpansion_unw = .
gen double Gap_DiD_unw     = .
gen double Expansion_wt    = .
gen double NoExpansion_wt  = .
gen double Gap_DiD_wt      = .

replace Expansion_unw   = real("`u12013'")    in 1
replace Expansion_unw   = real("`u12014'")    in 2
replace Expansion_unw   = real("`Trend_T_u'") in 3

replace NoExpansion_unw = real("`u02013'")    in 1
replace NoExpansion_unw = real("`u02014'")    in 2
replace NoExpansion_unw = real("`Trend_C_u'") in 3

replace Gap_DiD_unw     = real("`Gap13_u'")   in 1
replace Gap_DiD_unw     = real("`Gap14_u'")   in 2
replace Gap_DiD_unw     = real("`DiD_u'")     in 3

replace Expansion_wt    = real("`w12013'")    in 1
replace Expansion_wt    = real("`w12014'")    in 2
replace Expansion_wt    = real("`Trend_T_w'") in 3

replace NoExpansion_wt  = real("`w02013'")    in 1
replace NoExpansion_wt  = real("`w02014'")    in 2
replace NoExpansion_wt  = real("`Trend_C_w'") in 3

replace Gap_DiD_wt      = real("`Gap13_w'")   in 1
replace Gap_DiD_wt      = real("`Gap14_w'")   in 2
replace Gap_DiD_wt      = real("`DiD_w'")     in 3

format Expansion_unw NoExpansion_unw Gap_DiD_unw ///
       Expansion_wt  NoExpansion_wt  Gap_DiD_wt %9.1f

export excel using "../data/two_by_two_table.xlsx", firstrow(variables) replace

```
:::

## Table 3: Regression DiD

**Table 3** shows how we can recover the same $2\times2$ estimate using one of three regression specifications:

1\) Regressing the crude mortality rate on indicators for treatment (Medicaid expansion states), post-treatment periods (2014), and the interaction of the two.

2\) Regressing the crude mortality rate on the interaction of indicators for Medicaid expansion and post-expansion periods, as well as county and year fixed effects.

3\) Regressing the difference between 2014 and 2013 county-level crude mortality rates on an indicator for whether the county is in a state that expanded Medicaid.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
# show that you can get the same estimates with regression
# make a different short data with long differences in mortality by county and treatment indicators
short_data2 <- short_data %>% 
  group_by(county_code) %>% 
  summarize(state = state[1],
            set_wt = mean(set_wt),
            # long difference between 2014 and 2013 rates
            diff = crude_rate_20_64[which(year == 2014)] - crude_rate_20_64[which(year == 2013)],
            Treat = mean(Treat),
            Post = 1)

# estimate three models without weights. These are Treat*Post with no fixed effects,
# fixed effects + Treat:Post, and then the long difference model with no fixed effects.
mod1 <- feols(crude_rate_20_64 ~ Treat*Post, data = short_data, cluster = ~county_code)
mod2 <- feols(crude_rate_20_64 ~ Treat:Post | county_code + year, data = short_data, cluster = ~county_code)
mod3 <- feols(diff ~ Treat:Post, data = short_data2, cluster = ~county_code)

# estimate the same three three models with weights
mod4 <- feols(crude_rate_20_64 ~ Treat*Post, data = short_data, weights = ~set_wt, cluster = ~county_code)
mod5 <- feols(crude_rate_20_64 ~ Treat:Post | county_code + year, data = short_data, 
              weights = ~set_wt, cluster = ~county_code)
mod6 <- feols(diff ~ Treat:Post, data = short_data2, weights = ~set_wt, cluster = ~county_code)


# this dictionary maps names to labels for the table
dict <- c("(Intercept)" = "Constant",
          "Treat" = "Medicaid Expansion",
          "Post" = "Post",
          "Treat:Post" = "Medicaid Expansion X Post")

# rows to add to table
rows = tribble(
  ~x1, ~x2, ~x3, ~x4, ~x5, ~x6, ~x7,
  "County fixed effects", "No", "Yes", "No", "No", "Yes", "No",
  "Year fixed effects", "No", "Yes", "No", "No", "Yes", "No"
)

# export table 3
modelsummary(list("(1)" = mod1, "(2)" = mod2, "(3)" = mod3, 
                  "(4)" = mod4, "(5)" = mod5, "(6)" = mod6),
             coef_map = dict, escape = FALSE, gof_map = NA,
             fmt = 1, stars = FALSE,
             estimate = "{estimate}", add_rows = rows,
             title = "Regression 2 X 2 DiD") %>%
  add_header_above(c(" " = 1, "Crude Morality Rate" = 2, "\\(\\Delta\\)" = 1, 
                     "Crude Mortality Rate" = 2, "\\(\\Delta\\)" = 1), escape = FALSE)

```

### Stata

```{stata, eval = FALSE}

***************************************************************
*2. Table 3: regression 2x2 DiD with balanced panel, no controls
***************************************************************
	*Read in data, keep 2013–2014 and 2014 expansion vs "non expansion" states
	use "did_jel_aca_replication_data", clear
	keep if inlist(year, 2013, 2014) & inlist(yaca,2014,2020,2021,2023,.)
	gen Treat = (yaca==2014)		// 1 = ACA in 2014
	gen Post = (year == 2014)     	// 1 = post-expansion year

	*Panel declaration (once)
	xtset county_code year

	*UNWEIGHTED regressions
	eststo clear
	* (1) Treat × Post, no FE
	reg crude_rate_20_64 c.Treat##c.Post, vce(cluster county_code)
	eststo m1
	estadd local countyfe "No"
	estadd local yearfe   "No"

	* (2) Treat × Post with county & year FE  (xtreg, fe → no constant)
	xtreg crude_rate_20_64 c.Treat#c.Post i.year, fe vce(cluster county_code)
	eststo m2
	estadd local countyfe "Yes"
	estadd local yearfe   "Yes"

* (3) Long-difference, unweighted
preserve
    keep county_code year crude_rate_20_64 Treat
    reshape wide crude_rate_20_64, i(county_code) j(year)
    gen diff = crude_rate_20_642014 - crude_rate_20_642013
    label variable diff "\$\\Delta\$"
    reg diff Treat, vce(cluster county_code)

    // rename Treat -> c.Treat#c.Post and re-post BEFORE eststo
    tempname b V
    matrix `b' = e(b)
    matrix `V' = e(V)
    local N  = e(N)
    local df = e(df_r)
    tempvar touse
    gen byte `touse' = e(sample)
    local cn : colnames `b'
    local newcn
    foreach nm of local cn {
        if ("`nm'" == "Treat") local newcn `newcn' c.Treat#c.Post
        else                   local newcn `newcn' `nm'
    }
    matrix colnames `b' = `newcn'
    matrix colnames `V' = `newcn'
    matrix rownames `V' = `newcn'
    ereturn post `b' `V', depname(diff) esample(`touse') obs(`N') dof(`df')

    eststo m3
    estadd local countyfe "No"
    estadd local yearfe   "No"
restore

	*WEIGHTED regressions
	* (4) Weighted, no FE
	reg crude_rate_20_64 c.Treat##c.Post [aw=set_wt], vce(cluster county_code)
	eststo m4
	estadd local countyfe "No"
	estadd local yearfe   "No"

	* (5) Weighted, county & year FE  (areg keeps _cons; we'll hide it)
	xtreg crude_rate_20_64 c.Treat#c.Post i.year [aw=set_wt], ///
		fe vce(cluster county_code)
	eststo m5
	estadd local countyfe "Yes"
	estadd local yearfe   "Yes"


* (6) Weighted long-difference
preserve
    keep county_code year crude_rate_20_64 Treat set_wt
    reshape wide crude_rate_20_64, i(county_code) j(year)
    gen diff = crude_rate_20_642014 - crude_rate_20_642013
    reg diff Treat [aw=set_wt], vce(cluster county_code)

    // rename Treat -> c.Treat#c.Post and re-post BEFORE eststo
    tempname b V
    matrix `b' = e(b)
    matrix `V' = e(V)
    local N  = e(N)
    local df = e(df_r)
    tempvar touse
    gen byte `touse' = e(sample)
    local cn : colnames `b'
    local newcn
    foreach nm of local cn {
        if ("`nm'" == "Treat") local newcn `newcn' c.Treat#c.Post
        else                   local newcn `newcn' `nm'
    }
    matrix colnames `b' = `newcn'
    matrix colnames `V' = `newcn'
    matrix rownames `V' = `newcn'
    ereturn post `b' `V', depname(diff) esample(`touse') obs(`N') dof(`df')

    eststo m6
    estadd local countyfe "No"
    estadd local yearfe   "No"
restore
	

	*Build the LaTeX table
	local labels ///
		_cons            "Constant" ///
		Treat            "Medicaid Expansion" ///
		Post             "Post" ///
		c.Treat#c.Post   "Medicaid Expansion × Post"

	* Add Treat as the interaction label in long-diff specs (m3 and m6)
	estimates restore m3
	estadd local intlabel "Medicaid Expansion × Post"
	estimates restore m6
	estadd local intlabel "Medicaid Expansion × Post"

esttab m1 m2 m3 m4 m5 m6 using "../data/regdid_2x2.csv", replace keep(_cons Treat Post c.Treat#c.Post)   order(_cons Treat Post c.Treat#c.Post)  varlabels(_cons "Constant" Treat "Medicaid Expansion" Post  "Post"  c.Treat#c.Post "Medicaid Expansion × Post") cells(b(fmt(%9.1f) ) se(fmt(%9.1f) )) collabels(none) stats( countyfe yearfe, fmt(%9.0g %9s %9s) labels("County fixed effects" "Year fixed effects")) nomtitles nonotes noobs numbers  plain csv    


```
:::

## Table 4: Covariate Balance

In **Table 4** we show how to check for covariate balance across the covariates, using both 2013 levels and the changes between 2014 and 2013. We include a measure of the normalized difference as well as the (unweighted and weighted) sample means.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
# Now, make the covariate balance table (Table 4).
# unweighted - pre
unweighted <- short_data %>% 
  filter(year == 2013) %>% 
  select(Treat, all_of(covs)) %>% 
  group_by(Treat) %>% 
  # get mean and standard deviation
  summarize_all(list(mean, var)) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_fn1 - group0_fn1)/sqrt((group1_fn2 + group0_fn2)/2)) %>% 
  select(variable, group0_fn1, group1_fn1, norm_diff)

# make a weighted variance function
wtd.var <- function (x, weights = NULL, normwt = FALSE, na.rm = TRUE, 
                     method = c("unbiased", "ML")) 
{
  method <- match.arg(method)
  if (!length(weights)) {
    if (na.rm) 
      x <- x[!is.na(x)]
    return(var(x))
  }
  if (na.rm) {
    s <- !is.na(x + weights)
    x <- x[s]
    weights <- weights[s]
  }
  if (normwt) 
    weights <- weights * length(x)/sum(weights)
  if (normwt || method == "ML") 
    return(as.numeric(stats::cov.wt(cbind(x), weights, method = method)$cov))
  sw <- sum(weights)
  if (sw <= 1) 
    warning("only one effective observation; variance estimate undefined")
  xbar <- sum(weights * x)/sw
  sum(weights * ((x - xbar)^2))/(sw - 1)
}

# weighted  - pre
weighted <- short_data %>% 
  filter(year == 2013) %>% 
  select(Treat, all_of(covs), set_wt) %>% 
  group_by(Treat) %>% 
  # get mean and standard deviation
  summarize(across(all_of(covs), 
                   list(
                     ~weighted.mean(x = ., w = set_wt),
                     ~wtd.var(x = ., weights = set_wt, normwt = TRUE)))) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_1 - group0_1)/sqrt((group1_2 + group0_2)/2)) %>% 
  select(variable, group0_1, group1_1, norm_diff)

# make the top panel (weighted and unweighted Pre)
top_panel <- bind_cols(unweighted, weighted %>% select(-variable))

# make the bottom panel, which is the same thing but with the difference in X between 2014 and 2013.
# unweighted
unweighted <- short_data %>% 
  select(county_code, year, Treat, all_of(covs)) %>% 
  arrange(county_code, year) %>% 
  group_by(county_code, Treat) %>% 
  summarize(
    across(all_of(covs), function(x) x[2] - x[1])
  ) %>% 
  ungroup() %>% 
  select(-county_code) %>% 
  group_by(Treat) %>% 
  # get mean and standard deviation
  summarize_all(list(mean, var)) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_fn1 - group0_fn1)/sqrt((group1_fn2 + group0_fn2)/2)) %>% 
  select(variable, group0_fn1, group1_fn1, norm_diff)

# weighted
weighted <- short_data %>% 
  select(county_code, year, Treat, all_of(covs)) %>% 
  arrange(county_code, year) %>% 
  group_by(county_code, Treat) %>% 
  summarize(
    across(all_of(covs), function(x) x[2] - x[1])
  ) %>% 
  ungroup() %>% 
  left_join(short_data %>% filter(year == 2013) %>% select(county_code, set_wt),
            join_by(county_code)) %>% 
  select(-county_code) %>% 
  group_by(Treat) %>%
  # get mean and standard deviation
  summarize(across(all_of(covs), 
                   list(
                     ~weighted.mean(x = ., w = set_wt),
                     ~wtd.var(x = ., weights = set_wt, normwt = TRUE)))) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_1 - group0_1)/sqrt((group1_2 + group0_2)/2)) %>% 
  select(variable, group0_1, group1_1, norm_diff)

# make the bottom panel
bottom_panel <- bind_cols(unweighted, weighted %>% select(-variable))

# bind the two panels
table <- bind_rows(top_panel, bottom_panel) %>% 
  # reformat all columns to two digits
  mutate(across(-variable, \(x) scales::comma(x, accuracy = 0.01)))

# format Table 4 and export
table %>% 
  # change column names
  mutate(variable = 
           case_match(variable,
                      "perc_female" ~ "% Female",
                      "perc_white" ~ "% White",
                      "perc_hispanic" ~ "% Hispanic",
                      "unemp_rate" ~ "Unemployment Rate",
                      "poverty_rate" ~ "Poverty Rate",
                      "median_income" ~ "Median Income")) %>% 
  # format latex table
  kable(col.names = c("Variable", "Non-Adopt", "Adopt", "Norm. Diff.", "Non-Adopt", "Adopt", "Norm. Diff."),
        align = 'lcccccc',
        escape = T,
        booktabs = T,
        label = "cov_balance",
        caption = "Covariate Balance Statistics",
        linesep = "") %>% 
  kable_styling(latex_options = c("scale_down", "hold_position")) %>% 
  pack_rows("2013 Covariate Levels", 1, 6) %>% 
  pack_rows("2014 - 2013 Covariate Differences", 7, 12) %>% 
  add_header_above(c(" " = 1, "Unweighted" = 3, "Weighted" = 3))
```

### Stata

```{stata, eval = FALSE}
***************************************************************
*3. Table 4: Covariate-balance table
*			 NB:the table construction is a lot of lines here...
***************************************************************
	*Read in data, keep 2013–2014 and 2014 expansion vs "non expansion" states
	use "did_jel_aca_replication_data", clear
	keep if inlist(year, 2013, 2014) & inlist(yaca,2014,2020,2021,2023,.)
	gen Treat = (yaca==2014)		// 1 = ACA in 2014
	gen Post = (year == 2014)     	// 1 = post-expansion year

	*local with covariate names
	local covs perc_female perc_white perc_hispanic ///
			   unemp_rate_pc poverty_rate median_income_k

	*2013 levels — unweighted means, SDs, and normalized diff
	preserve
		keep if year == 2013

		*---- means and SDs in a single collapse -----------------------
		collapse (mean)  m_perc_female  			= perc_female       ///
						 m_perc_white   			= perc_white		///
						 m_perc_hispanic   			= perc_hispanic		///
						 m_unemp_rate_pc	  	 	= unemp_rate_pc 	///
						 m_poverty_rate 	     	= poverty_rate		///
						 m_median_income_k		 	= median_income_k 	///
				 (sd)    sd_perc_female  			= perc_female       ///
						 sd_perc_white   			= perc_white		///
						 sd_perc_hispanic  			= perc_hispanic		///
						 sd_unemp_rate_pc		   	= unemp_rate_pc 	///
						 sd_poverty_rate 	    	= poverty_rate		///
						 sd_median_income_k		 	= median_income_k, 	///
				 by(Treat)

		generate byte id = 1          // same value for both rows (Treat 0/1)

		*---- reshape into one wide row -------------------------------------
		reshape wide m* sd_*, i(id) j(Treat)
		drop id                       // no longer needed

		*---- normalized differences ----------------------------------
		foreach x in `covs'{
			gen nd_`x' = (m_`x'1 - m_`x'0) / ///
						 sqrt( (sd_`x'1^2 + sd_`x'0^2) / 2 )
		}

			keep m* nd_*                    // single-row dataset
			tempfile unwt_pre
			save `unwt_pre'
			list
	restore



	*2013 levels — weighted means, SDs, and normalized diff
	preserve
		keep if year == 2013

		*---- means and SDs in a single collapse -----------------------
		collapse (mean)  wm_perc_female  			= perc_female       ///
						 wm_perc_white   			= perc_white		///
						 wm_perc_hispanic   		= perc_hispanic		///
						 wm_unemp_rate_pc	  	 	= unemp_rate_pc 	///
						 wm_poverty_rate 	     	= poverty_rate		///
						 wm_median_income_k		 	= median_income_k 	///
				 (sd)    sd_perc_female  			= perc_female       ///
						 sd_perc_white   			= perc_white		///
						 sd_perc_hispanic  			= perc_hispanic		///
						 sd_unemp_rate_pc		   	= unemp_rate_pc 	///
						 sd_poverty_rate 	    	= poverty_rate		///
						 sd_median_income_k		 	= median_income_k 	///
						 [aw = set_wt], ///
				 by(Treat)

		generate byte id = 1          // same value for both rows (Treat 0/1)

		*---- reshape into one wide row -------------------------------------
		reshape wide wm* sd_*, i(id) j(Treat)
		drop id                       // no longer needed

		*---- normalized differences ----------------------------------
		foreach x in `covs'{
			gen wnd_`x' = (wm_`x'1 - wm_`x'0) / ///
						 sqrt( (sd_`x'1^2 + sd_`x'0^2) / 2 )
		}

			keep wm* wnd_*                    // single-row dataset
			tempfile wt_pre
			save `wt_pre'
			list
	restore


	* Differences - unweighted means, SDs, and normalized diff 
	preserve
		*create differences
		for var `covs': egen bX = total(X*(year==2013)), by(county_code)
		for var `covs': gen d_X = X-bX
    
	  keep if year == 2014 

		*---- means and SDs in a single collapse -----------------------
		collapse (mean)  m_d_perc_female  			= d_perc_female         ///
						 m_d_perc_white   			= d_perc_white			///
						 m_d_perc_hispanic   		= d_perc_hispanic		///
						 m_d_unemp_rate_pc	  	 	= d_unemp_rate_pc 		///
						 m_d_poverty_rate 	     	= d_poverty_rate		///
						 m_d_median_income_k	 	= d_median_income_k 	///
				 (sd)    sd_perc_female  			= d_perc_female         ///
						 sd_perc_white   			= d_perc_white			///
						 sd_perc_hispanic  			= d_perc_hispanic		///
						 sd_unemp_rate_pc	   		= d_unemp_rate_pc 		///
						 sd_poverty_rate 	    	= d_poverty_rate		///
						 sd_median_income_k		 	= d_median_income_k, 	///
				 by(Treat)

		generate byte id = 1          // same value for both rows (Treat 0/1)

		*---- reshape into one wide row -------------------------------------
		reshape wide m* sd_*, i(id) j(Treat)
		drop id                       // no longer needed

		*---- normalized differences ----------------------------------
		foreach x in `covs'{
			gen nd_d_`x' = (m_d_`x'1 - m_d_`x'0) / ///
						 sqrt( (sd_`x'1^2 + sd_`x'0^2) / 2 )
		}

		keep m_d* nd_d*                    // single-row dataset
		tempfile unwt_diff
		save `unwt_diff'
		list	
	restore

	* Differences - weighted means, SDs, and normalized diff 
	preserve
		*create differences
		for var `covs': egen bX = total(X*(year==2013)), by(county_code)
		for var `covs': gen d_X = X-bX
    keep if year == 2014 

		*---- means and SDs in a single collapse -----------------------
		collapse (mean)  wm_d_perc_female  			= d_perc_female         ///
						 wm_d_perc_white   			= d_perc_white			///
						 wm_d_perc_hispanic   		= d_perc_hispanic		///
						 wm_d_unemp_rate_pc	  	 	= d_unemp_rate_pc 		///
						 wm_d_poverty_rate 	     	= d_poverty_rate		///
						 wm_d_median_income_k	 	= d_median_income_k 	///
				 (sd)    sd_perc_female  			= d_perc_female         ///
						 sd_perc_white   			= d_perc_white			///
						 sd_perc_hispanic  			= d_perc_hispanic		///
						 sd_unemp_rate_pc	   		= d_unemp_rate_pc 		///
						 sd_poverty_rate 	    	= d_poverty_rate		///
						 sd_median_income_k		 	= d_median_income_k 	///
						 [aw = set_wt], ///
				 by(Treat)

		generate byte id = 1          // same value for both rows (Treat 0/1)

		*---- reshape into one wide row -------------------------------------
		reshape wide wm* sd_*, i(id) j(Treat)
		drop id                       // no longer needed

		*---- normalized differences ----------------------------------
		foreach x in `covs'{
			gen wnd_d_`x' = (wm_d_`x'1 - wm_d_`x'0) / ///
						 sqrt( (sd_`x'1^2 + sd_`x'0^2) / 2 )
		}

		keep wm* wnd_*                    // single-row dataset
		tempfile wt_diff
		save `wt_diff'
		list	
	restore

	**table creation
	*Assemble the four blocks into one matrix
	use `unwt_pre',     clear
	merge 1:1 _n using `wt_pre',   nogen   // adds mw*  sw*  wnd_*
	merge 1:1 _n using `unwt_diff', nogen   // adds md*  nd_d*
	merge 1:1 _n using `wt_diff',  nogen   // adds mwd* swd* wnd_d_*

	de,f

		   
	* ---- row labels ----------------------------------------------------
	local rowlbl  "% Female" "% White" "% Hispanic" ///
				  "Unemployment Rate" "Poverty Rate" "Median Income"
	local outrows
	local i = 1
	foreach v in `covs' {
		local outrows `"`outrows' `"`:word `i' of `rowlbl''"'"'
		local ++i
	}

	/* ---- build the 12 × 6 matrix ----------------------------------- */
	matrix T = J(12,6,.)
	local r 0
	foreach v of local covs {
		local ++r
		* rows 1–6 : 2013 levels
		matrix T[`r',1] = m_`v'0[1]         // un-wtd mean, Non-adopt
		matrix T[`r',2] = m_`v'1[1]         // un-wtd mean, Adopt
		matrix T[`r',3] = nd_`v'[1]         // un-wtd norm-diff
		matrix T[`r',4] = wm_`v'0[1]        // wtd mean,  Non-adopt
		matrix T[`r',5] = wm_`v'1[1]        // wtd mean,  Adopt
		matrix T[`r',6] = wnd_`v'[1]        // wtd norm-diff
	}

	foreach v of local covs {
		local ++r
		* rows 7–12 : 2014–2013 long-differences
		matrix T[`r',1] = m_d_`v'0[1]        // un-wtd ∆, Non-adopt
		matrix T[`r',2] = m_d_`v'1[1]        // un-wtd ∆, Adopt
		matrix T[`r',3] = nd_d_`v'[1]        // un-wtd norm-diff
		matrix T[`r',4] = wm_d_`v'0[1]       // **underscore added**
		matrix T[`r',5] = wm_d_`v'1[1]       // **underscore added**
		matrix T[`r',6] = wnd_d_`v'[1]       // wtd norm-diff
	}

	matrix rownames T = `outrows' `outrows'
	matrix colnames T = "Non-Adopt" "Adopt" "Norm. Diff." ///
						"Non-Adopt" "Adopt" "Norm. Diff."
						

				
*--------------------------------------------------------------------
* 5.  Export to CSV instead of LaTeX
*--------------------------------------------------------------------

* Convert matrix T to a dataset
clear
svmat T

* Reorder so variable names come first

* Rename columns to something clear
rename T1 NonAdopt_unwt
rename T2 Adopt_unwt
rename T3 NormDiff_unwt
rename T4 NonAdopt_wt
rename T5 Adopt_wt
rename T6 NormDiff_wt

* After svmat/renames
foreach v of varlist NonAdopt_unwt Adopt_unwt NormDiff_unwt ///
                     NonAdopt_wt  Adopt_wt  NormDiff_wt {
    replace `v' = round(`v', .01)
	  tostring `v', replace format(%9.2f) force
                     }


* Save as CSV
export delimited using "../data/cov_balance.csv", replace

					
```
:::

## Table 5: Regression 2 $\times$ 2 DiD with Covariates

In **Table 5** we show how the 2 $\times$ 2 DiD estimates change with the inclusion of covariates using linear regression. You can do this in two ways - either controlling for the 2013 levels of the covariates, or the change in the covariates between 2014 and 2013. Even in this simple setting the results differ considerably.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
# make a table that includes the model 1) without covariates, 2) long regression with 
# 2013 covariates values, 3) long regression with *difference* in covariate values.
reg_data_2013 <- short_data %>% 
  # make long diff in y
  group_by(county_code) %>% 
  summarize(long_y = crude_rate_20_64[which(year == 2014)] - crude_rate_20_64[which(year == 2013)]) %>% 
  # merge in 2013 covariates
  left_join(short_data %>% filter(year == 2013) %>% select(county_code, state, Treat, set_wt, all_of(covs)), 
            by = "county_code")

# this dataset has the change in X between 2014 and 2013 as controls
reg_data_change <- short_data %>% 
  # make long diff in y
  group_by(county_code) %>% 
  summarize(long_y = crude_rate_20_64[which(year == 2014)] - crude_rate_20_64[which(year == 2013)]) %>% 
  # merge in change in covariate values
  left_join(short_data %>% 
              group_by(county_code) %>% 
              mutate(set_wt = set_wt[which(year == 2013)]) %>% 
              group_by(county_code, state, Treat, set_wt) %>% 
              summarize(
                across(all_of(covs), function(x) x[which(year == 2014)] - x[which(year == 2013)])
              ), by = "county_code")

# run the six models
# first unweighted - long diff no covs, 2013 covs, then change in covs
mod1 <- feols(long_y ~ Treat, data = reg_data_2013, cluster = ~county_code)
mod2 <- feols(long_y ~ Treat + .[covs], data = reg_data_2013, cluster = ~county_code)
mod3 <- feols(long_y ~ Treat + .[covs], data = reg_data_change, cluster = ~county_code)

# same thing but weighted
mod4 <- feols(long_y ~ Treat, data = reg_data_2013, weights = ~set_wt, cluster = ~county_code)
mod5 <- feols(long_y ~ Treat + .[covs], data = reg_data_2013, weights = ~set_wt, cluster = ~county_code)
mod6 <- feols(long_y ~ Treat + .[covs], data = reg_data_change, weights = ~set_wt, cluster = ~county_code)

# export table 3
modelsummary(list("(1)" = mod1, "(2)" = mod2, "(3)" = mod3, 
                  "(4)" = mod4, "(5)" = mod5, "(6)" = mod6),
             coef_map = dict, coef_omit = "(Intercept)", escape = FALSE, gof_map = NA,
             fmt = 2, estimate = "{estimate}",
             title = "Regression 2 X 2 DiD with Covariates") %>%
  add_header_above(c(" " = 1, "No Covs" = 1, "$$X_{i, t = 2013}$$" = 1, "$$X_{i, t}$$" = 1,
                                    "No Covs" = 1, "$$X_{i, t = 2013}$$" = 1, "$$X_{i, t}$$" = 1), 
                   escape = FALSE, extra_css = 'vertical-align: middle !important;', line = F) %>% 
  add_header_above(c(" " = 1, "Unweighted" = 3, "Weighted" = 3), escape = FALSE)
  
```

### Stata

```{stata, eval = FALSE}

***************************************************************
*4. Table 5: Regression 2 x 2 DiD with Covariates 
***************************************************************
		*read data in again and restrict to 2013/2014
		use "did_jel_aca_replication_data", clear
		keep if inlist(year, 2013, 2014) & inlist(yaca,2014,2020,2021,2023,.)
		gen Treat = (yaca==2014)		// 1 = ACA in 2014
		gen Post = (year == 2014)     	// 1 = post-expansion year

		*local with covariate names
		local covs perc_female perc_white perc_hispanic ///
			   unemp_rate_pc poverty_rate median_income_k

		* Keep only relevant variables
		keep county_code year Treat set_wt crude_rate_20_64 `covs'

		* Reshape everything (outcome + covariates) at once
		reshape wide crude_rate_20_64 `covs', i(county_code) j(year)

		* Generate change in covariates
		foreach v of local covs {
			gen d_`v' = `v'2014 - `v'2013
		}

		* Generate outcome: change in crude mortality
		gen long_y = crude_rate_20_642014 - crude_rate_20_642013

		* Drop missing outcome
		drop if missing(long_y)


	eststo clear

	* (1) Unweighted: No covariates
	reg long_y Treat, vce(cluster county_code)
	eststo m1

	* (2) Unweighted: 2013 covariates
	reg long_y Treat ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013, ///
		vce(cluster county_code)
	eststo m2

	* (3) Unweighted: Changes in covariates
	reg long_y Treat ///
		d_perc_female d_perc_white d_perc_hispanic ///
		d_unemp_rate_pc d_poverty_rate d_median_income_k, ///
		vce(cluster county_code)
	eststo m3

	* (4) Weighted: No covariates
	reg long_y Treat [aw=set_wt], vce(cluster county_code)
	eststo m4

	* (5) Weighted: 2013 covariates
	reg long_y Treat ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013 ///
		[aw=set_wt], vce(cluster county_code)
	eststo m5

	* (6) Weighted: Changes in covariates
	reg long_y Treat ///
		d_perc_female d_perc_white d_perc_hispanic ///
		d_unemp_rate_pc d_poverty_rate d_median_income_k ///
		[aw=set_wt], vce(cluster county_code)
	eststo m6

esttab m1 m2 m3 m4 m5 m6 using "../data/regdid_2x2_covs.csv", replace keep(Treat) b(%6.2f) se(%6.2f) nostar fragment plain csv label nomtitles alignment(c) collabels(none) 
				 

```
:::

## Table 6: Outcome Regression and Propensity Score Models

We next show how to adjust for covariates in the 2 $\times$ 2 setting in a disciplined manner. **Table 6** reports the outcome regression and propensity score model results that feed into the doubly-robust estimator from Callaway and Sant'Anna (2021).

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
## Next we make Table 6 which shows the pscore and outcome models that feed into CS.
# These are done by regressing long y on the covariates for the untreated units (the outcome model)
# and regressing the expansion indicator on the covariates for 2013 data (the propensity model) 
# We do it with and without weights
mod1 <- feols(long_y ~ .[covs], data = reg_data_2013 %>% filter(Treat == 0), cluster = ~county_code)
mod2 <- feglm(Treat ~ .[covs], data = short_data %>% filter(year == 2013), family = "binomial", vcov = "hetero")
mod3 <- feols(long_y ~ .[covs], data = reg_data_2013 %>% filter(Treat == 0), cluster = ~county_code, weights = ~set_wt)
mod4 <- feglm(Treat ~ .[covs], data = short_data %>% filter(year == 2013), family = "binomial", vcov = "hetero", weights = ~set_wt)

# This is the dictionary to map variables to data labels for the table
dict <- c("(Intercept)" = "Constant", 
          "perc_female" = "% Female",
          "perc_white" = "% White",
          "perc_hispanic" = "% Hispanic",
          "crude_rate_20_64" = "Crude Mortality Rate",
          "unemp_rate" = "Unemployment Rate",
          "poverty_rate" = "Poverty Rate",
          "median_income" = "Median Income")

# export table 3
modelsummary(list("(1)" = mod1, "(2)" = mod2, "(3)" = mod3, "(4)" = mod4),
             coef_map = dict, escape = FALSE, gof_map = NA,
             fmt = 2, estimate = "{estimate}",
             title = "Outcome Regression and Propensity Score Models") %>%
  add_header_above(c(" " = 1, "Regression" = 1, "Propensity Score" = 1, 
                     "Regression" = 1, "Propensity Score" = 1), 
                   escape = FALSE, extra_css = 'vertical-align: middle !important;', line = F) %>% 
  add_header_above(c(" " = 1, "Unweighted" = 2, "Weighted" = 2), escape = FALSE)

```

### Stata

```{stata, eval = FALSE}
		 
***************************************************************
* Table 6: Outcome Regressions + Propensity Score Models
***************************************************************
	*read data in again and restrict to 2013/2014
	use "did_jel_aca_replication_data", clear
	keep if inlist(year, 2013, 2014) & inlist(yaca,2014,2020,2021,2023,.)
	gen Treat = (yaca==2014)		// 1 = ACA in 2014
	gen Post = (year == 2014)     	// 1 = post-expansion year

	*local with covariate names
	local covs perc_female perc_white perc_hispanic ///
		   unemp_rate_pc poverty_rate median_income_k

	* Keep only relevant variables
	keep county_code year Treat set_wt crude_rate_20_64 `covs'

	* Reshape everything (outcome + covariates) at once
	reshape wide crude_rate_20_64 `covs', i(county_code) j(year)

	* Generate change in covariates
	foreach v of local covs {
		gen d_`v' = `v'2014 - `v'2013
	}

	* Generate outcome: change in crude mortality
	gen long_y = crude_rate_20_642014 - crude_rate_20_642013

	* Drop missing outcome
	drop if missing(long_y)
	
	eststo clear

	** (1) Outcome Regression — Unweighted, Untreated Only
	reg long_y ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013 if Treat==0, ///
		vce(cluster county_code)
	eststo or1

	** (2) Propensity Score Model — Unweighted Logit
	logit Treat ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013, ///
		vce(cluster county_code)
	eststo ps1

	** (3) Outcome Regression — Weighted, Untreated Only
	reg long_y ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013 if Treat==0 ///
		[aw = set_wt], vce(cluster county_code)
	eststo or2

	** (4) Propensity Score — Weighted Logit
	logit Treat ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013 ///
		[iw = set_wt], vce(cluster county_code)
	eststo ps2

	esttab or1 ps1 or2 ps2 using "../data/reg_pscore.csv", replace /// 
    cells(b(fmt(%9.2f) ) se(fmt(%9.2f) )) nostar collabels(none)  ///
    fragment label alignment(c) order(_cons ) nomtitles nonotes noobs numbers  plain csv

```
:::

## Table 7: Sant'Anna and Zhao (2020) DiD

**Table 7** shows how to calculate the DiD estimate using the estimators in Sant'Anna and Zhao (2020). The paper describes three different estimation methods: outcome regression, inverse-probability weighted (IPW), and doubly robust models.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
# finally, get the same estimates using Sant'Anna and Zhao (2020) and Callaway and Sant'Anna (2021) using the 
# three adjustment methods
# You need to reformat the group variable (untreated = 0) and the unit ID variable needs to be numeric
data_cs <- short_data %>% 
  mutate(treat_year = if_else(yaca == 2014 & !is.na(yaca), 2014, 0),
         county_code = as.numeric(county_code))

# create a function to run the CS estimator allowing the 
# estimation method to differ
run_cs <- function(method, wt) {
  
  # estimate the att_gt
  atts <- att_gt(
    yname = "crude_rate_20_64",
    tname = "year",
    idname = "county_code",
    gname = "treat_year",
    xformla =  as.formula(paste("~", paste(covs, collapse = "+"))),
    data = data_cs,
    panel = TRUE,
    control_group = "nevertreated",
    bstrap = TRUE,
    cband = TRUE,
    est_method = method,
    weightsname = wt,
    # faster_mode = TRUE,
    base_period = "universal",
    biters = 25000
  )
  
  # aggregate estimates 
  aggte(atts, na.rm = TRUE, biters = 25000) %>% 
    broom::tidy() %>% 
    filter(group == 2014) %>% 
    mutate(type = method)
  
}

# run it three ways - regression adjustment, IPW, and doubly robust
# these are unweighted
ests <- map_dfr(c("reg", "ipw", "dr"), run_cs, wt = NULL)

# make a table - Panel A is with unweighted models
tablea <- ests %>% 
  # reformat the data
  select(type, estimate, std.error) %>% 
  # format the estimate and std error
  mutate(estimate = scales::number(estimate, accuracy = 0.01),
         std.error = paste0("(", scales::number(std.error, accuracy = 0.01), ")")) %>% 
  # reshape the data
  pivot_longer(cols = -type,
               names_to = "statistic",
               values_to = "value") %>% 
  pivot_wider(id_cols = "statistic",
              names_from = "type",
              values_from = "value") %>% 
  mutate(statistic = if_else(statistic == "estimate", "Medicaid Expansion", " "))

# make weighted results for the three CS models.
ests_w <-  map_dfr(c("reg", "ipw", "dr"), run_cs, wt = "set_wt")

# Panel B is the same thing but with weighted results
tableb <- ests_w %>% 
  select(type, estimate, std.error) %>% 
  mutate(estimate = scales::number(estimate, accuracy = 0.01),
         std.error = paste0("(", scales::number(std.error, accuracy = 0.01), ")")) %>% 
  pivot_longer(cols = -type,
               names_to = "statistic",
               values_to = "value") %>% 
  pivot_wider(id_cols = "statistic",
              names_from = "type",
              values_from = "value") %>% 
  select(reg, ipw, dr)

# Combine Panels A and B together and format the table
bind_cols(tablea, tableb) %>% 
  kable(col.names = c(" ", "Regression", "IPW", "Doubly Robust",
                      "Regression", "IPW", "Doubly Robust"),
        align = 'lcccccc',
        escape = F,
        booktabs = T,
        label = "2x2_csdid",
        caption = "Callaway and Sant'Anna (2021) DiD",
        linesep = "") %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  add_header_above(c(" " = 1, "Unweighted" = 3, "Weighted" = 3))
```

### Stata

```{stata, eval = FALSE}
		 
***************************************************************
* Table 7: 2x2 DiD with covariates (Sant'Anna and Zhao 2020)
***************************************************************
	*read data in again and restrict to 2013/2014
	use "did_jel_aca_replication_data", clear
	keep if inlist(year, 2013, 2014) & inlist(yaca,2014,2020,2021,2023,.)
	gen Treat = (yaca==2014)		// 1 = ACA in 2014
	gen Post = (year == 2014)     	// 1 = post-expansion year

	*local with covariate names
	local covs perc_female perc_white perc_hispanic ///
		   unemp_rate_pc poverty_rate median_income_k

	*create baseline covariates (drdid would do this but we do it explicitly here)
	*drop the time-varying version of the covariate; not used here
	foreach x in `covs'{
		egen `x'2013 =total(`x'*(year==2013)),by(county_code)
		drop `x'
	}
	* rename the baseline covariates just created 
	rename (perc_female2013 perc_white2013 perc_hispanic2013 ///
			unemp_rate_pc2013 poverty_rate2013 median_income_k2013) ///
		   (perc_female perc_white perc_hispanic ///
			unemp_rate_pc poverty_rate median_income_k)

	* Set panel
	xtset county_code year

	* Unweighted DRDID Estimation
	*request all estimators in one shot (we only use reg, stdipw, and dripw)
	drdid crude_rate_20_64 `covs', ///
		  ivar(county_code) time(year) tr(Treat) ///
		  all ///
		  pscoretrim(0.995) ///
		  wboot(reps(25000) rseed(20240924) wbtype(rademacher)) ///
		  cluster(county_code) 

	* Extract point estimates
	scalar reg     = e(b)[1,3]
	scalar ipw     = e(b)[1,5]
	scalar dripw   = e(b)[1,1]

	* Extract standard errors
	scalar se_reg   = sqrt(e(V)[3,3])
	scalar se_ipw   = sqrt(e(V)[5,5])
	scalar se_dripw = sqrt(e(V)[1,1])

	* Format for LaTeX export
	local reg_str     : display %6.2f reg
	local ipw_str     : display %6.2f ipw
	local dripw_str   : display %6.2f dripw

	local se_reg_str     : display %6.2f se_reg
	local se_ipw_str     : display %6.2f se_ipw
	local se_dripw_str   : display %6.2f se_dripw


	* Weighted DRDID Estimation
	*request all estimators in one shot (we only use reg, stdipw, and dripw)
	drdid crude_rate_20_64 `covs' [iweight = set_wt], ///
		  ivar(county_code) time(year) tr(Treat) ///
		  all ///
		  pscoretrim(0.995) ///
		  wboot(reps(25000) rseed(20240924) wbtype(rademacher)) ///
		  cluster(county_code) 
	
		
* Extract point estimates
scalar reg_w     = e(b)[1,3]
scalar ipw_w     = e(b)[1,5]
scalar dripw_w   = e(b)[1,1]

* Extract standard errors
scalar se_reg_w   = sqrt(e(V)[3,3])
scalar se_ipw_w   = sqrt(e(V)[5,5])
scalar se_dripw_w = sqrt(e(V)[1,1])

*Make matrix of the results of interest 
matrix results_wide = ( ///
    reg    , ipw    , dripw    , reg_w    , ipw_w    , dripw_w   \ ///
    se_reg , se_ipw , se_dripw , se_reg_w , se_ipw_w , se_dripw_w )

matrix rownames results_wide = b se
matrix colnames results_wide = Unw_REG Unw_IPW Unw_DRIPW W_REG W_IPW W_DRIPW

matrix list results_wide
clear
svmat results_wide, names(col)

* add row labels b/se
gen stat = ""
replace stat = "Medicaid Expansion"  in 1
replace stat = "" in 2

order stat Unw_REG Unw_IPW Unw_DRIPW W_REG W_IPW W_DRIPW
export delimited using "../data/reg_t7.csv", replace




```
:::

## Figure 1: Distribution of Propensity Scores

**Figure 1** plots the distribution of the propensity scores from the underlying propensity model for the unweighted and weighted models.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10
# Finally, we make Figure 1 which reports the distribution of the propensity scores between 
# expansion and non-expansion counties.
# first add in propensity scores to the data
plot_data <- bind_rows(
  short_data %>% 
    filter(year == 2013) %>% 
    mutate(propensity = predict(mod2, ., type = "response"),
           mod = "Unweighted",
           wt = 1),
  short_data %>% 
    filter(year == 2013) %>% 
    mutate(propensity = predict(mod4, ., type = "response"),
           mod = "Weighted", 
           wt = set_wt)
  )

# plot the propensity score distributions by group
plot_data %>% 
  mutate(expand = if_else(Treat == 1, "Expansion Counties", "Non-Expansion Counties")) %>% 
  ggplot(aes(x = propensity, y = after_stat(density), group = expand, color = expand, weight = wt)) + 
  geom_histogram(fill = "white", position = "identity", linewidth = 1, alpha = 0.5) + 
  facet_wrap(~mod) + 
  scale_color_brewer(palette = 'Set1') + 
  labs(x = "Propensity Score", y = "Density") + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

### Stata

```{stata, eval = FALSE}

*************************************************
** Figure 1: Distribution of Propensity Scores
*************************************************
	*read data in again and restrict to 2013/2014
	use "did_jel_aca_replication_data", clear
	keep if inlist(year, 2013, 2014) & inlist(yaca,2014,2020,2021,2023,.)
	gen Treat = (yaca==2014)		// 1 = ACA in 2014
	gen Post = (year == 2014)     	// 1 = post-expansion year

	*local with covariate names
	local covs perc_female perc_white perc_hispanic ///
		   unemp_rate_pc poverty_rate median_income_k

	* Keep only relevant variables
	keep county_code year Treat set_wt crude_rate_20_64 `covs'

	* Reshape everything (outcome + covariates) at once
	reshape wide crude_rate_20_64 `covs', i(county_code) j(year)

	* Generate change in covariates
	foreach v of local covs {
		gen d_`v' = `v'2014 - `v'2013
	}

	* Generate outcome: change in crude mortality
	gen long_y = crude_rate_20_642014 - crude_rate_20_642013

	* Drop missing outcome
	drop if missing(long_y)
	
	
	*Unweighted p-score distributions
	eststo ps1: logit Treat ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013, ///
		vce(cluster county_code)
	
	
	*estimates use ps1
	predict pscore_unw

	* Expansion counties
	summ pscore_unw if inlist(Treat,0,1)
	local min = r(min)
	local max = r(max)
	local w   = (`max' - `min')/30
	local s   = floor(`min'/`w')*`w'   // align start to a bin edge
	di `w'
	di `s'
	
	twoway 	histogram pscore_unw if Treat == 1 , width(`w') start(`s') fcolor(none) lcolor(red) || ///
			histogram pscore_unw if Treat == 0 , width(`w') start(`s') fcolor(none) lcolor(blue) ///
			ytitle("Density",angle(horizontal)) xtitle("") title("Unweighted") legend(off) ylabel(0(1)4.5) xlabel(,nogrid)
			
	graph save "pscore_uw.gph", replace
	

	*Weighted p-score distributions
	eststo ps2: logit Treat ///
		perc_female2013 perc_white2013 perc_hispanic2013 ///
		unemp_rate_pc2013 poverty_rate2013 median_income_k2013, ///
		[iw = set_wt], vce(cluster county_code)
	
	
	*estimates use ps2
	predict pscore_w

	* Define group for legend
	gen group = cond(Treat == 1, "Expansion", "Non-Expansion")

	* Expansion counties
	summ pscore_w if inlist(Treat,0,1)
	local min = r(min)
	local max = r(max)
	local w   = (`max' - `min')/30
	local s   = floor(`min'/`w')*`w'   // align start to a bin edge

	twoway 	histogram pscore_w if Treat == 1 [fw=set_wt], width(`w') start(`s') fcolor(none) lcolor(red) || ///
			histogram pscore_w if Treat == 0 [fw=set_wt], width(`w') start(`s') fcolor(none) lcolor(blue) ///
			ytitle("") xtitle("Propensity Score") title("Weighted") legend( rows(1)) ylabel(0(1)4.5) xlabel(, nogrid)

	graph save "pscore_w.gph", replace

	*combine graphs and use only one legend and y-axis title
	**NB: differences between the Stata and R histograms are purely down to 
	*some arcane disagreement between histogram and ggplot; as can be seen from 
	*the actual p-score model estimates, the pscores themselves are identical
	grc1leg2 pscore_uw.gph pscore_w.gph, ///
		title("") xtob1title xtitlefrom("pscore_w.gph") ///
		legendfrom(pscore_w.gph) position(6) ring(2) ///
		cols(2) imargin(4 4 4 4)
	graph display, xsize(10) ysize(5)




```

:::

# 2 $\times$ T DiD setups

In the next section we show how to estimate DiD when you have more than two time periods, what we call the 2 $\times$ T setting.

## Figure 2: County Mortality Trends by Expansion Decision

In **Figure 2** we plot the weighted average county mortality rates over our sample period (2009-2019) for our two groups: those who expand in 2014 and those who have not expanded by 2019.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10
# set seed for reproducibility
set.seed(20240924)

# load cleaned data
mydata <- read_csv(here::here("data", "county_mortality_data.csv")) %>% 
  # make state the abbreviation
  mutate(state = str_sub(county, nchar(county) - 1, nchar(county))) %>% 
  # drop DC and pre-2014 adoption states
  filter(!(state %in% c("DC", "DE", "MA", "NY", "VT"))) %>% 
  # drop states that adopt between 2014 and 2019
  filter(yaca == 2014 | is.na(yaca) | yaca > 2019)

# set the covariates that we're going to use.
covs <- c("perc_female","perc_white", "perc_hispanic", 
          "unemp_rate", "poverty_rate", "median_income")

## Clean data and add in covariates
mydata <- mydata %>% 
  # make variables
  mutate(perc_white = population_20_64_white / population_20_64 * 100,
         perc_hispanic = population_20_64_hispanic / population_20_64 * 100,
         perc_female = population_20_64_female/ population_20_64 * 100,
         unemp_rate = unemp_rate * 100,
         median_income = median_income / 1000) %>% 
  # keep just subset of variables that we will use later
  select(state, county, county_code, year, population_20_64, yaca,
         starts_with("perc_"), crude_rate_20_64, all_of(covs))

# keep only counties with full observations for outcome and covariates in 2013 and 2014
mydata <- mydata %>%
  # allow the aca expansion variable to be missing
  drop_na(!yaca) %>%
  group_by(county_code) %>% 
  # need full covariates for 2013 and 2014
  filter(length(which(year == 2013 | year == 2014)) == 2) %>% 
  ungroup()

# finally, keep only counties with full mortality data for 2009 to 2019
mydata <- mydata %>% 
  group_by(county_code) %>% 
  drop_na(crude_rate_20_64) %>% 
  filter(n() == 11)

# make ACA and post variables, as well as the weighting variable which is the relevant county 
# population in 2013
mydata <- mydata %>% 
  mutate(Treat = if_else(yaca == 2014 & !is.na(yaca), 1, 0),
         Post = if_else(year >= 2014, 1, 0)) %>% 
  group_by(county_code) %>% 
  # make a variable with population weight in 2013
  mutate(set_wt = population_20_64[which(year == 2013)]) %>% 
  ungroup()

# make time series plot
mydata %>% 
  mutate(expand = if_else(Treat == 1, "Expansion Counties", "Non-Expansion Counties")) %>% 
  # get the weighted averages by expansion type and year
  group_by(expand, year) %>% 
  summarize(mortality = weighted.mean(crude_rate_20_64, set_wt)) %>% 
  # plot
  ggplot(aes(x = year, y = mortality, group = expand, color = expand)) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  geom_vline(xintercept = 2014, linetype = "dashed") + 
  scale_color_brewer(palette = 'Set1') + 
  scale_x_continuous(breaks = 2009:2019) + 
  labs(x = "", y = "Mortality (20-64) \n Per 100,000") + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

### Stata

```{stata, eval = FALSE}

***************************************************************
*1. Figure 2: 2014 versus non-expansion mortality trends
***************************************************************	
	*read data in again and restrict to 2013/2014
	use "did_jel_aca_replication_data", clear
	keep if inlist(yaca,2014,2020,2021,2023,.)
	gen Treat = (yaca==2014)		// 1 = ACA in 2014

	*Collapse to means by group × year
	gen expand = cond(Treat==1, "Expansion Counties", "Non-Expansion Counties")

	collapse (mean) crude_rate_20_64 [iw=set_wt], by(expand year)
	rename crude_rate_20_64 mortality

	* Create the plot
	twoway ///
	  (line mortality year if expand == "Expansion Counties",      lcolor(red)  lpattern(solid) lwidth(medium)) ///
	  (line mortality year if expand == "Non-Expansion Counties",  lcolor(blue) lpattern(solid)  lwidth(medium)) ///
	  (scatter mortality year if expand == "Expansion Counties",   msymbol(O)   mcolor(red)) ///
	  (scatter mortality year if expand == "Non-Expansion Counties", msymbol(Oh) mcolor(blue))  , ///
	  legend(order(1 "Expansion Counties" 2 "Non-Expansion Counties") ///
			 pos(6) ring(2) cols(2)) ///
	  xtitle("") ///
	  ytitle("Mortality (20–64)" "Per 100,000", orientation(horizontal)) ///
	  ysize(5) xsize(10) ///
	  xline(2013) ///
	  xlabel(2009(1)2019, nogrid) ///
	  graphregion(color(white)) ///
	  plotregion(style(none))





```

:::

## Figure 3: 2 $\times$ T Event Study

In **Figure 3** we report the $2 \times T$ event study without covariates.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10
# Make event study plots
# first add in the variables needed to run the CS model - timing group variable (treat_year)
# and county code needs to be numeric
mydata <- mydata %>% 
  mutate(treat_year = if_else(yaca == 2014 & !is.na(yaca), 2014, 0),
         county_code = as.numeric(county_code),
         time_to_treat = if_else(Treat == 1, year - treat_year, 0))

# estimate CS models

# get the individual ATT(g,t) estimates
mod <- did::att_gt(
  yname = "crude_rate_20_64",
  tname = "year",
  idname = "county_code",
  gname = "treat_year",
  xformla =  NULL,
  data = mydata,
  panel = TRUE,
  control_group = "nevertreated",
  bstrap = TRUE,
  cband = TRUE,
  est_method = "reg",
  weightsname = "set_wt",
  # faster_mode = TRUE,
  base_period = "universal",
  biters = 25000
)

# confirm you get the same with OLS (standard errors differ because of bootstrap only)
cs_out <- mod

ols_out <- feols(crude_rate_20_64 ~ i(time_to_treat, Treat, ref = -1) | county_code + year, 
      data = mydata, 
      cluster = ~county_code, 
      weights = ~set_wt)

# get the Rambachan/Roth confidence interval
# first save our aggregate event study estimates
es <- did::aggte(
  mod,
  type = "dynamic",
  min_e = -5,
  max_e = 0,
  bstrap = TRUE,
  biters = 25000
)

# source the honest_did code made to fit our estimates
source(here::here('scripts/R/5_honestdid.R'))

# get robust CI
robust_ci <- honest_did(es = es, type = "relative_magnitude")$robust_ci

# get the aggregate value for e = 0:5
agg <- mod %>% 
  aggte(type = "dynamic", min_e = 0, max_e = 5,
        bstrap = TRUE, biters = 25000)

# make labels for the estimate, std error, and confidence intervals
label1 <-  paste0("Estimate~(e %in% '{0, 5}')~'='~'", scales::number(agg$overall.att, accuracy = 0.01), "'")

label2 <- paste0("Std. Error = ", scales::number(agg$overall.se, 0.01), " \n",
                 "Conf. Int = [", scales::number(agg$overall.att - 1.96*agg$overall.se, 0.01), ", ", 
                 scales::number(agg$overall.att + 1.96*agg$overall.se, 0.01), "]")

# make event study plot for Figure 3
mod %>% 
  # Aggregate in event time ("dynamic")
  aggte(type = "dynamic", biters = 25000) %>% 
  # get the two confidence intervals
  broom::tidy(conf.int = TRUE) %>% 
  # plot
  ggplot(aes(x = event.time, y = estimate)) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high), color = "darkred") + 
  geom_linerange(aes(ymin = point.conf.low, ymax = point.conf.high)) + 
  geom_point() + 
  geom_vline(xintercept = -1, linetype = "dashed") + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  scale_x_continuous(breaks = -5:5) + 
  annotate("text", x = 3, y = 11, label = label1, parse = TRUE) + 
  annotate("text", x = 3, y = 9, label = label2) + 
  labs(x = "Event Time", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

### Stata

```{stata, eval = FALSE}



***************************************************************
*2. Figure 3: 2xT event-study with no covariates
***************************************************************	
	*read data in again and restrict to 2013/2014
	use "did_jel_aca_replication_data", clear
	keep if inlist(yaca,2014,2020,2021,2023,.)

	*Use csdid for simple ES estimation (uniform and non-uniform CIs); use long2!!
	csdid crude_rate_20_64 [iw=set_wt], ///
		ivar(county_code) time(year) gvar(treat_year) ///
		long2 ///
		wboot(reps(25000) rseed(20240924) wbtype(rademacher)) ///
		agg(event)
		
	*you do actually have to grab the uniform CIs from the posted output
	preserve
		clear
		matrix T  = r(table)
		local cn : colnames T
		matrix llu = T[5,1..colsof(T)]
		matrix ulu = T[6,1..colsof(T)]
		matrix uci = (llu', ulu')
		matrix colnames uci = uci_lower uci_upper
		svmat uci, names(col)
		gen var = ""
		local i = 0
		foreach r of local cn{
			local ++i
			replace var = "`r'" in `i'
		}
		save "testo", replace
	restore

	*save the posted output to a dataset and merge back on the uniform CIs
	regsave, ci
	merge 1:1 var using "testo"
	keep if _merge==3
	drop _merge
	
	*get average post coefficient, standard error, and CI
	qui sum coef if var=="Post_avg"
	local postcoef : display %03.2f r(mean)
	qui sum stderr if var=="Post_avg"
	local postse : display %03.2f r(mean)
	local postlci : display %03.2f `postcoef'-1.96*`postse'
	local postuci : display %03.2f `postcoef'+1.96*`postse'
	
	*now only keep the event-study estimates
	keep if substr(var,1,1)=="T"
	keep coef ci* uci* var
	
	*extract timing components from event aggregation notation in csdid
	gen pre = substr(var,2,1)=="m"
	gen post = substr(var,2,1)=="p"
	gen e = real(substr(var,3,.))
	replace e = -e if pre
	
	*add omitted period
	local obs = _N+1
	set obs `obs'
	replace coef = 0 in `obs'
	replace e = -1 in `obs'


	*plot the ATT(g,t) for g=2014
	twoway ///
		(rcap uci_lower uci_upper e, ///
		lcolor(red*1.2) msize(0)) ///
		(rcap ci_lower ci_upper e, ///
		lcolor(black) msize(0)) ///
		(scatter coef e, ///
		msym(O) mcolor(black) 	///
		), ///
		legend(off) ///
		xlabel(-5(1)5, nogrid) ///
		ylabel(-5(5)10) ///	
		yline(0, lpattern(dash)) ///
		xline(-1, lpattern(dash)) ///
		xtitle("Event-Time") ///
		ytitle("Treatment Effect" "Mortality Per 100,000", orientation(horizontal) size(small)) ///
		text(10.5 3 "Estimate e ∈ {0,5}=`postcoef'", size(small)) ///
		text(9 3 "Std. Error = `postse'" "Conf. Int = [`postlci',`postuci']", size(small))	///
		xsize(10) ysize(5)
	



```

:::

## Rambachan & Roth (2023) Sensitivity Analysis

In the text surrounding the $2 \times T$ event study, we also discuss how you can use the analysis developed in Rambachan and Roth (2023) to test the sensitivity of the ATT estimates to violations of the parallel trends assumption. In particular, we focus on their "relative magnitude" analysis, wherein we assume that any parallel trends violations in the post-period are no larger than the largest violation in the pre-period. In the code below we show how to estimate the credible interval under this bounding approach for $ATT(2014)$ - or the treatment effect in the first year of treatment for those counties that adopted Medicaid expansion.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10

# get robust CI
robust_ci %>% 
  # keep just relative magnitude = 1
  filter(Mbar == 1) %>%
  # drop extraneous variables
  select(lb, ub, Mbar) %>% 
  # round
  mutate_at(c("lb", "ub"), scales::number, accuracy = 0.01) %>% 
  # make table
  kable(col.names = c("lower_bound", "upper_bound", "relative_magnitude"),
       align = 'c', booktabs = T) %>% 
  kable_styling()

```

### Stata

```{stata, eval = FALSE}
************************************************
*1. read data in again and restrict to 2013/2014
************************************************
	use "did_jel_aca_replication_data", clear
	keep if inlist(yaca,2014,2020,2021,2023,.)
	
************************************************
*Estimate unconditional csdid version 1.81 (Oct 2025)
************************************************
	csdid crude_rate_20_64 [iw=set_wt], ///
		ivar(county_code) time(year) gvar(treat_year) ///
		long2 ///
		never ///
		wboot(reps(25000) rseed(20240924) wbtype(rademacher)) ///
		agg(event)

	estat event, window(-5,0) post
	
************************************************
*Honest DiD, relative magnitude only using M=1 for e=0
************************************************	
honestdid, type(relative_magnitude) pre(3/6) post(7) mvec(1)

exit

```
:::

## Figure 4: $2 \times T$ Event Study With Covariates

In **Figure 4** we show the $2 \times T$ event studies with covariates using the Callaway and Sant'Anna (2021) estimators - i.e. using outcome regression, IPW, and doubly-robust estimators.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10
# Estimate the CS Model 3 ways with the three estimation types
# make a function to run CS, varying the estimation type
run_cs <- function(method) {
  
  # get the ATT(g, t) estimates
  att_gt(
    yname = "crude_rate_20_64",
    tname = "year",
    idname = "county_code",
    gname = "treat_year",
    xformla =  as.formula(paste("~", paste(covs, collapse = "+"))),
    data = mydata,
    panel = TRUE,
    control_group = "nevertreated",
    bstrap = TRUE,
    cband = TRUE,
    est_method = method,
    weightsname = "set_wt",
    # faster_mode = TRUE,
    base_period = "universal", 
    biters = 25000
  ) %>% 
    # aggregate to event time (dynamic)
    aggte(type = "dynamic", na.rm = TRUE, biters = 25000) %>% 
    broom::tidy(conf.int = TRUE) %>% 
    mutate(type = method)
}

# estimate it for 3 methods 
out <- map_dfr(c("reg", "ipw", "dr"), run_cs)

# make plot for Figure 4 - 2XT Event Study with Covariates
out %>% 
  # refactor labels for plot
  mutate(type = case_match(type,
                           "reg" ~ "Regression",
                           "ipw" ~ "IPW",
                           "dr" ~ "Doubly Robust")) %>% 
  mutate(type = factor(type, levels = c("Regression", "IPW", "Doubly Robust"))) %>% 
  # plot the esti mates and confidence intervals
  ggplot(aes(x = event.time, y = estimate)) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high), color = "darkred") + 
  geom_linerange(aes(ymin = point.conf.low, ymax = point.conf.high)) + 
  geom_point() + 
  geom_vline(xintercept = -1, linetype = "dashed") + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  facet_wrap(~type) + 
  scale_x_continuous(breaks = -5:5) + 
  labs(x = "Event Time", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

```

### Stata

```{stata, eval = FALSE}


***************************************************************
*3. Figure 4: 2xT event-study with covariates
***************************************************************	
	*read data in again and restrict to 2013/2014
	use "did_jel_aca_replication_data", clear
	keep if inlist(yaca,2014,2020,2021,2023,.)
	
	*local with covariate names
	local covs perc_female perc_white perc_hispanic ///
		   unemp_rate_pc poverty_rate median_income_k

	*Estimate RA, IPW, and DR using csdid version 1.81 (Oct 2025)
	local reg_title "Regression"
	local stdipw_title "IPW"
	local dripw_title "Doubly Robust"
	foreach m in reg stdipw dripw{
		csdid crude_rate_20_64 `covs' [iw=set_wt], ///
			ivar(county_code) time(year) gvar(treat_year) ///
			method(`m') ///
			long2 ///
			never ///
			wboot(reps(25000) rseed(20240924) wbtype(rademacher)) ///
			pscoretrim(0.995) agg(event)

		*you do actually have to grab the uniform CIs from the posted output
		preserve
			clear
			matrix T  = r(table)
			local cn : colnames T
			matrix llu = T[5,1..colsof(T)]
			matrix ulu = T[6,1..colsof(T)]
			matrix uci = (llu', ulu')
			matrix colnames uci = uci_lower uci_upper
			svmat uci, names(col)
			gen var = ""
			local i = 0
			foreach r of local cn{
				local ++i
				replace var = "`r'" in `i'
			}
			save "testo", replace
		restore

		*save the posted output to a dataset and merge back on the uniform CIs
		preserve
			regsave, ci
			merge 1:1 var using "testo"
			keep if _merge==3
			drop _merge
			
			*get average post coefficient, standard error, and CI
			qui sum coef if var=="Post_avg"
			local postcoef : display %03.2f r(mean)
			qui sum stderr if var=="Post_avg"
			local postse : display %03.2f r(mean)
			local postlci : display %03.2f `postcoef'-1.96*`postse'
			local postuci : display %03.2f `postcoef'+1.96*`postse'
			
			*now only keep the event-study estimates
			keep if substr(var,1,1)=="T"
			keep coef ci* uci* var
			
			*extract timing components from event aggregation notation in csdid
			gen pre = substr(var,2,1)=="m"
			gen post = substr(var,2,1)=="p"
			gen e = real(substr(var,3,.))
			replace e = -e if pre
			
			*add omitted period
			local obs = _N+1
			set obs `obs'
			replace coef = 0 in `obs'
			replace e = -1 in `obs'


			*plot the ATT(g,t) for g=2014
			twoway ///
				(rcap uci_lower uci_upper e, ///
				lcolor(red*1.2) msize(0)) ///
				(rcap ci_lower ci_upper e, ///
				lcolor(black) msize(0)) ///
				(scatter coef e, ///
				msym(O) mcolor(black) 	///
				), ///
				legend(off) ///
				xlabel(-5(1)5, nogrid) ///
				ylabel(-20(10)20) ///	
				xline(-1, lpattern(dash)) ///
				yline(0, lpattern(dash)) ///
				xtitle("Event-Time") ///
				ytitle("") ///
				title("``m'_title'") ///
				xsize(10) ysize(5)
				
				graph save "`m'.gph", replace
		restore
	}

	*combine graphs and use only one legend and y-axis title
	**NB: differences between the Stata and R come from the bootstrap uncertainty
	**analytic standard errors are identical
	grc1leg2 reg.gph stdipw.gph dripw.gph, ///
		title("") xtob1title xtitlefrom("reg.gph") ///
		l1title("Treatment Effect" "Mortality Per 100,000", size(small) orientation(horizontal)) ///
		loff ///
		ycommon ///
		position(6) ring(2) ///
		cols(3) imargin(4 4 4 4)

	graph display, xsize(12) ysize(5)




exit






```

:::

# G $\times$ T DiD setups

We end the paper with the staggered adoption setting, with $G$ treatment groups and $T > 2$ time periods.

## Figure 5: Mortality Trends by Expansion Decision with Staggered Timing

In **Figure 5** we plot the trends in weighted-average county mortality rates by treatment timing groups from 2009 to 2019.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10

# set seed for reproducibility
set.seed(20240924)

# load cleaned data
mydata <- read_csv(here::here("data", "county_mortality_data.csv")) %>% 
  # make state the abbreviation
  mutate(state = str_sub(county, nchar(county) - 1, nchar(county))) %>% 
  # drop DC and pre-2014 adoption states
  filter(!(state %in% c("DC", "DE", "MA", "NY", "VT"))) 

# set the covariates that we're going to use.
covs <- c("perc_female","perc_white", "perc_hispanic", "unemp_rate", "poverty_rate", "median_income")

## Clean data and add in covariates
mydata <- mydata %>% 
  # make variables
  mutate(perc_white = population_20_64_white / population_20_64 * 100,
         perc_hispanic = population_20_64_hispanic / population_20_64 * 100,
         perc_female = population_20_64_female/ population_20_64 * 100,
         unemp_rate = unemp_rate * 100,
         median_income = median_income / 1000) %>% 
  # keep just subset of variables that we will use later
  select(state, county, county_code, year, population_20_64, yaca,
         starts_with("perc_"), crude_rate_20_64, all_of(covs))

# keep only counties with full observations for outcome and covariates in 2013 and 2014
mydata <- mydata %>%
  # allow the aca expansion variable to be missing
  drop_na(!yaca) %>%
  group_by(county_code) %>% 
  # need full covariates for 2013 and 2014
  filter(length(which(year == 2013 | year == 2014)) == 2) %>% 
  ungroup()

# finally, keep only counties with full mortality data for 2009 to 2019
mydata <- mydata %>% 
  group_by(county_code) %>% 
  drop_na(crude_rate_20_64) %>% 
  filter(n() == 11)

# make smaller dataset
mydata <- mydata %>% 
  mutate(Treat = if_else(!is.na(yaca) & yaca <= 2019, 1, 0),
         treat_year = if_else(!is.na(yaca) & yaca <= 2019, yaca, 0),
         Post = if_else(year >= 2014, 1, 0)) %>% 
  group_by(county_code) %>% 
  # make a variable with population weight in 2013
  mutate(set_wt = population_20_64[which(year == 2013)]) %>% 
  ungroup()

# make time series plot by timing group (Figure 5)
mydata %>% 
  # identify the groups
  mutate(treat_year = if_else(treat_year == 0, "Non-Expansion Counties", as.character(treat_year))) %>% 
  # get weighted average mortality by gtiming-group and year
  group_by(treat_year, year) %>% 
  summarize(mortality = weighted.mean(crude_rate_20_64, set_wt)) %>% 
  ggplot(aes(x = year, y = mortality, group = as.factor(treat_year), color = as.factor(treat_year))) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  scale_color_manual(values = c("#7C7189", "#D04E59", "#BC8E7D", "#2F3D70", "#CABEE9")) + 
  scale_x_continuous(breaks = 2009:2019) + 
  labs(x = "", y = "Mortality (20-64) \n Per 100,000") + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

### Stata

```{stata, eval = FALSE}

**************************************************************************
*1. Figure 5: Mortality Trends by Timing Group
**************************************************************************
	*read data in again and relabel treatment groups, specifically the "non-expansion" group
	use "did_jel_aca_replication_data", clear
	
	*Define treatment timing labels
	replace treat_str = "Non-Expansion Counties" if inlist(treat_str, "0", "2020", "2021", "2022", "2023")

	collapse (mean) crude_rate [aw = set_wt], by(treat_str year)

	* Plot trends by Timing Group
	twoway ///
		(connected crude_rate year if treat_str == "Non-Expansion Counties", ///
			lcolor(purple*0.35) lwidth(medthick) mcolor(purple*0.35) msymbol(O)) ///
		(connected crude_rate year if treat_str == "2014", ///
			lcolor(gray*1.2) lwidth(medthick) mcolor(gray*1.2) msymbol(O)) ///
		(connected crude_rate year if treat_str == "2015", ///
			lcolor(red) lwidth(medthick) mcolor(red*1.5) msymbol(O)) ///
		(connected crude_rate year if treat_str == "2016", ///
			lcolor(brown) lwidth(medthick) mcolor(brown) msymbol(O)) ///
		(connected crude_rate year if treat_str == "2019", ///
			lcolor(navy) lwidth(medthick) mcolor(navy) msymbol(O)), ///
		///
		legend(order(2 "2014" 3 "2015" 4 "2016" 5 "2019" 1 "Non-Expansion Counties") ///
				pos(6) ring(2) rows(1) region(lcolor(none)) size(small)) ///
		ytitle("Mortality (20–64)" "per 100,000", size(small) orientation(horizontal)) ///
		ylabel(350 400 450, angle(0) labsize(small) nogrid) ///
		xtitle("") xlabel(2009(1)2019, labsize(small)) ///
		graphregion(color(white)) ///
		name(fig5, replace)

		graph display, xsize(10) ysize(5)




```

:::

## Figure 6: ATT(g, t)s for Each Expansion Group

With the estimator from Callaway & Sant'Anna (2021) we can estimate the group-time average treatment effects - what we call ATT(g, t) - for each group and time period. We plot these cohort-specific treatment effects in **Figure 6**.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10
# reclassify the county code as numeric for Callaway/Sant'Anna
mydata <- mydata %>% 
  mutate(county_code = as.numeric(county_code))

# get the ATT(g,t)'s
mod_no_x <- att_gt(
  yname = "crude_rate_20_64",
  tname = "year",
  idname = "county_code",
  gname = "treat_year",
  xformla =  NULL,
  data = mydata,
  panel = TRUE,
  control_group = "notyettreated",
  bstrap = TRUE,
  biters = 25000,
  cband = TRUE,
  weightsname = "set_wt",
  # faster_mode = TRUE,
  base_period = "universal"
)

# make plots in calendar time by timing group
mod_no_x %>% 
  # get the estimates and confidence intervals
  broom::tidy(conf.int = TRUE) %>% 
  mutate(group = as.character(group)) %>% 
  # plot the estimates and CIs
  ggplot(aes(x = time, y = estimate, color = group)) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high)) + 
  scale_x_continuous(breaks = seq(2009, 2019, by = 2),
                     labels = seq(2009, 2019, by = 2)) + 
  scale_color_manual(values = c("#7C7189", "#D04E59", "#BC8E7D", "#2F3D70")) + 
  geom_vline(aes(xintercept = as.numeric(group) - 1), linetype = "dashed", linewidth = 1) + 
  labs(x = "", y = "Treatment Effect \n Mortality Per 100,000") + 
  facet_wrap(~group) + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

### Stata

```{stata, eval = FALSE}

**************************************************************************
*2.   Figure 6: ATT(g,t) by Calendar Time
**************************************************************************
	*read data in 
	use "did_jel_aca_replication_data", clear

	*Estimate group × time effects
	csdid crude_rate [iw=set_wt], ///
		ivar(county_code) ///
		time(year) ///
		gvar(treat_year) ///
		long2 ///
		notyet ///
		wboot(reps(25000) rseed(20240924) wbtype(rademacher)) 
	
	*adding the omitted category to every g is very annoying ni the matrix style, so do the graphs using regsave	
	regsave
	
	*extract timing components from csdid
	gen g = real(substr(var,2,4))
	gen t1 = real(substr(var,9,4))	
	gen t2 = real(substr(var,14,4))	
	gen t = (t1)*(t2==g-1) + (t2)*(t1==g-1)
	
	*add omitted period 
	preserve
		keep g
		duplicates drop
		gen t = g-1
		gen coef = 0
		save testo, replace
	restore
	append using testo
	
	*make CIs
	gen lci = coef - 1.96*stderr
	gen uci = coef + 1.96*stderr	
	keep coef lci uci g t
	sort g t
	
	*plot the ATT(g,t) for g=2014
	twoway ///
		(rcap lci uci t if g==2014, ///
		lcolor(gray*1.2) msize(0)) ///
		(scatter coef t if g==2014, ///
		msym(O) mcolor(gray*1.2) 	///
		c(l) lcolor(gray*1.2) 		///
		), ///
		legend(off) ///
		xlabel(2009(2)2019, nogrid) ///
		ylabel(-40(20)20) ///	
		xline(2013, lpattern(dash)) ///
		title("2014") ///
		xtitle("") ///
		ytitle("", orientation(horizontal)) ///
		xsize(10) ysize(5)
		
		graph save g2014.gph, replace

	*plot the ATT(g,t) for g=2015
	twoway ///
		(rcap lci uci t if g==2015, ///
		lcolor(red*1.5) msize(0)) ///
		(scatter coef t if g==2015, ///
		msym(O) mcolor(red*1.5) 	///
		c(l) lcolor(red*1.5) 		///
		), ///
		legend(off) ///
		xlabel(2009(2)2019, nogrid) ///
		ylabel(-40(20)20) ///	
		xline(2014, lpattern(dash)) ///
		title("2015") ///
		xtitle("") ///
		ytitle("", orientation(horizontal)) ///
		xsize(10) ysize(5)
		
		graph save g2015.gph, replace

	*plot the ATT(g,t) for g=2016
	twoway ///
		(rcap lci uci t if g==2016, ///
		lcolor(brown) msize(0)) ///
		(scatter coef t if g==2016, ///
		msym(O) mcolor(brown) 	///
		c(l) lcolor(brown) 		///
		), ///
		legend(off) ///
		xlabel(2009(2)2019, nogrid) ///
		ylabel(-40(20)20) ///	
		xline(2015, lpattern(dash)) ///
		title("2016") ///
		xtitle("") ///
		ytitle("", orientation(horizontal)) ///
		xsize(10) ysize(5)		
		
		graph save g2016.gph, replace
		
	*plot the ATT(g,t) for g=2019
	twoway ///
		(rcap lci uci t if g==2019, ///
		lcolor(navy) msize(0)) ///
		(scatter coef t if g==2019, ///
		msym(O) mcolor(navy) 	///
		c(l) lcolor(navy) 		///
		), ///
		legend(off) ///
		xlabel(2009(2)2019, nogrid) ///
		ylabel(-40(20)20) ///	
		xline(2018, lpattern(dash)) ///
		title("2019") ///
		xtitle("") ///
		ytitle() ///
		xsize(10) ysize(5)		
		
		graph save g2019.gph, replace
				
	graph combine g2014.gph g2015.gph g2016.gph g2019.gph, ///
		title("Treatment Effect" "Mortality per 100,000", size(small) orientation(horizontal) pos(9) ring(2))  ///
		cols(2) imargin(4 4 4 4)  
		
	graph display, xsize(10) ysize(5)

		


```

:::

## Figure 7: ATT(g, t) in Event Time

We show how you can stack these group-time average treatment effects in relative, or event, time in **Figure 7**.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE}
#| fig.align: 'center'
#| fig.width: 10

# Make a different plot that reports the ATT(g,t)s in relative time
mod_no_x %>% 
  # get the estimates and CIs
  broom::tidy() %>% 
  # define relative time for each group
  mutate(rel_time = time - group) %>% 
  # plot
  ggplot(aes(x = rel_time, y = estimate, group = as.factor(group), color = as.factor(group))) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  geom_vline(xintercept = -1, linetype = "dashed", linewidth = 1) + 
  scale_color_manual(values = c("#7C7189", "#D04E59", "#BC8E7D", "#2F3D70")) + 
  scale_x_continuous(breaks = -10:5) + 
  labs(x = "", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

### Stata

```{stata, eval = FALSE}


**************************************************************************
*3.   Figure 7: ATT(g,t) in event-time
**************************************************************************
		*read data in 
	use "did_jel_aca_replication_data", clear

	*Estimate group × time effects
	csdid crude_rate [iw=set_wt], ///
		ivar(county_code) ///
		time(year) ///
		gvar(treat_year) ///
		long2 ///
		notyet ///
		wboot(reps(25000) rseed(20240924) wbtype(rademacher)) 
	
	*adding the omitted category to every g is very annoying ni the matrix style, so do the graphs using regsave	
	regsave
	
	*extract timing components from csdid
	gen g = real(substr(var,2,4))
	gen t1 = real(substr(var,9,4))	
	gen t2 = real(substr(var,14,4))	
	gen t = (t1)*(t2==g-1) + (t2)*(t1==g-1)
	
	*add omitted period 
	preserve
		keep g
		duplicates drop
		gen t = g-1
		gen coef = 0
		save testo, replace
	restore
	append using testo
	
	*make CIs
	gen lci = coef - 1.96*stderr
	gen uci = coef + 1.96*stderr	
	keep coef lci uci g t
	sort g t
  
  gen e = t-g
	keep g e coef
	keep if g<.
	reshape wide coef, i(e) j(g)
	twoway ///
		scatter coef2014 coef2015 coef2016 coef2019 e, ///
		msym(o o o o) mcolor(gray*1.2 red*1.5 brown navy) ///
		c(l l l l) lcolor(gray*1.2 red*1.5 brown navy) lwidth(medthick medthick medthick medthick) ///
		legend(order(1 "2014" 2 "2015" 3 "2016" 4 "2019") rows(1) pos(6) ring(2) size(small)) ///
		xlabel(-10(1)5, nogrid) ///
		ylabel(-20(10)20) ///	
		xline(-1, lpattern(dash)) ///
		title("") ///
		xtitle("") ///
		ytitle() ///
		xsize(10) ysize(5)	
	



```

:::

## Figure 8: G $\times$ T Event Study Without Covariates

Finally, we demonstrate how to estimate DiD in the $G \times T$ setting by leverage treatment timing and not-yet-treated groups for identification using the doubly-robust method from Callaway and Sant'Anna (2021). First, in **Figure 8** we show the DiD event study estimates without covariates.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE, collectcode=TRUE}
#| fig.align: 'center'
#| fig.width: 10
# Make a full -5 to + 5 event study plot for no covariates 
# using Callaway/Sant'Anna with not-yet-treated

# get the aggregate value for e = 0:5
agg <- mod_no_x %>% 
  aggte(type = "dynamic", min_e = 0, max_e = 5,
        bstrap = TRUE, biters = 25000)

# make labels for the estimate, std error, and confidence intervals
label1 <-  paste0("Estimate~(e %in% '{0, 5}')~'='~'", scales::number(agg$overall.att, accuracy = 0.01), "'")

label2 <- paste0("Std. Error = ", scales::number(agg$overall.se, 0.01), " \n",
                 "Conf. Int = [", scales::number(agg$overall.att - 1.96*agg$overall.se, 0.01), ", ", 
                 scales::number(agg$overall.att + 1.96*agg$overall.se, 0.01), "]")

# make G*T event study plot without covariates
mod_no_x %>% 
  # aggregate into relative time (dynamic)
  aggte(type = "dynamic") %>% 
  broom::tidy(conf.int = TRUE) %>% 
  # keep just -5 to + 5
  filter(event.time %>% between(-5, 5)) %>% 
  # plot
  ggplot(aes(x = event.time, y = estimate)) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high), color = "darkred") + 
  geom_linerange(aes(ymin = point.conf.low, ymax = point.conf.high)) + 
  geom_point() + 
  geom_vline(xintercept = -1, linetype = "dashed") + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  scale_x_continuous(breaks = -5:5) + 
  annotate("text", x = 3, y = 11, label = label1, parse = TRUE) + 
  annotate("text", x = 3, y = 9, label = label2) + 
  labs(x = "Event Time", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

```

### Stata

```{stata, eval = FALSE}


/**************************************************************************
4. Figure 8: Event study without covariates
**************************************************************************/
	*read data in 
	use "did_jel_aca_replication_data", clear

	*Use csdid for simple ES estimation (uniform and non-uniform CIs); use long2!!
	csdid crude_rate_20_64 [iw=set_wt], ///
		ivar(county_code) time(year) gvar(treat_year) ///
		long2 ///
		notyet ///
		wboot(reps(25000) rseed(20240924) wbtype(rademacher)) ///
		agg(event)

	*you do actually have to grab the uniform CIs from the posted output
	preserve
		clear
		matrix T  = r(table)
		local cn : colnames T
		matrix llu = T[5,1..colsof(T)]
		matrix ulu = T[6,1..colsof(T)]
		matrix uci = (llu', ulu')
		matrix colnames uci = uci_lower uci_upper
		svmat uci, names(col)
		gen var = ""
		local i = 0
		foreach r of local cn{
			local ++i
			replace var = "`r'" in `i'
		}
		save "testo", replace
	restore

	*save the posted output to a dataset and merge back on the uniform CIs
	regsave, ci
	merge 1:1 var using "testo"
	keep if _merge==3
	drop _merge
	
	*get average post coefficient, standard error, and CI
	qui sum coef if var=="Post_avg"
	local postcoef : display %03.2f r(mean)
	qui sum stderr if var=="Post_avg"
	local postse : display %03.2f r(mean)
	local postlci : display %03.2f `postcoef'-1.96*`postse'
	local postuci : display %03.2f `postcoef'+1.96*`postse'
	
	*now only keep the event-study estimates
	keep if substr(var,1,1)=="T"
	keep coef ci* uci* var
	
	*extract timing components from event aggregation notation in csdid
	gen pre = substr(var,2,1)=="m"
	gen post = substr(var,2,1)=="p"
	gen e = real(substr(var,3,.))
	replace e = -e if pre
	
	*add omitted period
	local obs = _N+1
	set obs `obs'
	replace coef = 0 in `obs'
	replace e = -1 in `obs'

	*plot only -5 to 5
	keep if inrange(e,-5,5)

	*plot the ATT(g,t) for g=2014
	twoway ///
		(rcap uci_lower uci_upper e, ///
		lcolor(red*1.2) msize(0)) ///
		(rcap ci_lower ci_upper e, ///
		lcolor(black) msize(0)) ///
		(scatter coef e, ///
		msym(O) mcolor(black) 	///
		), ///
		legend(off) ///
		xlabel(-5(1)5, nogrid) ///
		ylabel(-20(10)20) ///	
		yline(0, lpattern(dash)) ///
		xline(-1, lpattern(dash)) ///
		xtitle("Event-Time") ///
		ytitle("Treatment Effect" "Mortality Per 100,000", orientation(horizontal) size(small)) ///
		text(18.5 3 "Estimate e ∈ {0,5}=`postcoef'", size(small)) ///
		text(15.5 3 "Std. Error = `postse'" "Conf. Int = [`postlci',`postuci']", size(small))	///
		xsize(10) ysize(5)





```

:::

## Figure 9: G $\times$ T Event Study With Covariates

**Figure 9** presents the event study results adjusting for the covariates using the doubly-robust method.

::: panel-tabset
### R

```{r, message = FALSE, warning = FALSE, error = FALSE, collectcode=TRUE}
#| fig.align: 'center'
#| fig.width: 10
# Now re-estimate and plot the same GXT estimates with covariates
# set.seed(20240924)
mod_with_x <- att_gt(
  yname = "crude_rate_20_64",
  tname = "year",
  idname = "county_code",
  gname = "treat_year",
  xformla =  as.formula(paste("~", paste(covs, collapse = "+"))),
  data = mydata,
  panel = TRUE,
  control_group = "notyettreated",
  bstrap = TRUE,
  biters = 25000,
  cband = TRUE,
  est_method = "dr",
  weightsname = "set_wt",
  # faster_mode = TRUE,
  base_period = "universal"

)

# get the aggregate value for e = 0:5
agg <- mod_with_x %>% 
  aggte(type = "dynamic", min_e = 0, max_e = 5,
        bstrap = TRUE, biters = 25000)

# make labels for the estimate, std error, and confidence intervals
label1 <-  paste0("Estimate~(e %in% '{0, 5}')~'='~'", scales::number(agg$overall.att, accuracy = 0.01), "'")

label2 <- paste0("Std. Error = ", scales::number(agg$overall.se, 0.01), " \n",
                 "Conf. Int = [", scales::number(agg$overall.att - 1.96*agg$overall.se, 0.01), ", ", 
                 scales::number(agg$overall.att + 1.96*agg$overall.se, 0.01), "]")

# make event study plot for GXT with covariates
mod_with_x %>% 
  # aggregate to event time (dynamic)
  aggte(type = "dynamic") %>% 
  broom::tidy(conf.int = TRUE) %>% 
  # filter years
  filter(event.time %>% between(-5, 5)) %>% 
  # plot
  ggplot(aes(x = event.time, y = estimate)) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high), color = "darkred") + 
  geom_linerange(aes(ymin = point.conf.low, ymax = point.conf.high)) + 
  geom_point() + 
  geom_vline(xintercept = -1, linetype = "dashed") + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  scale_x_continuous(breaks = -5:5) + 
  annotate("text", x = 3, y = 18, label = label1, parse = TRUE) + 
  annotate("text", x = 3, y = 15, label = label2) + 
  labs(x = "Event Time", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

```

### Stata

```{stata, eval = FALSE}


/**************************************************************************
5. Figure 9: Event study with covariates
**************************************************************************/
	*read data in 
	use "did_jel_aca_replication_data", clear

	*local with covariate names
	local covs perc_female perc_white perc_hispanic ///
		   unemp_rate_pc poverty_rate median_income_k
	
	*Use csdid for simple ES estimation (uniform and non-uniform CIs); use long2!!
	csdid crude_rate_20_64 `covs' [iw=set_wt], ///
		ivar(county_code) time(year) gvar(treat_year) ///
		method(dripw) ///
		long2 ///
		notyet ///
		pscoretrim(0.995) ///
		wboot(reps(25000) rseed(20240924) wbtype(rademacher)) 	///
		agg(event)

	*you do actually have to grab the uniform CIs from the posted output
	preserve
		clear
		matrix T  = r(table)
		local cn : colnames T
		matrix llu = T[5,1..colsof(T)]
		matrix ulu = T[6,1..colsof(T)]
		matrix uci = (llu', ulu')
		matrix colnames uci = uci_lower uci_upper
		svmat uci, names(col)
		gen var = ""
		local i = 0
		foreach r of local cn{
			local ++i
			replace var = "`r'" in `i'
		}
		save "testo", replace
	restore

	*save the posted output to a dataset and merge back on the uniform CIs
	regsave, ci
	merge 1:1 var using "testo"
	keep if _merge==3
	drop _merge
	
	*get average post coefficient, standard error, and CI
	qui sum coef if var=="Post_avg"
	local postcoef : display %03.2f r(mean)
	qui sum stderr if var=="Post_avg"
	local postse : display %03.2f r(mean)
	local postlci : display %03.2f `postcoef'-1.96*`postse'
	local postuci : display %03.2f `postcoef'+1.96*`postse'
	
	*now only keep the event-study estimates
	keep if substr(var,1,1)=="T"
	keep coef ci* uci* var
	
	*extract timing components from event aggregation notation in csdid
	gen pre = substr(var,2,1)=="m"
	gen post = substr(var,2,1)=="p"
	gen e = real(substr(var,3,.))
	replace e = -e if pre
	
	*add omitted period
	local obs = _N+1
	set obs `obs'
	replace coef = 0 in `obs'
	replace e = -1 in `obs'

	*plot only -5 to 5
	keep if inrange(e,-5,5)

	*plot the ATT(g,t) for g=2014
	twoway ///
		(rcap uci_lower uci_upper e, ///
		lcolor(red*1.2) msize(0)) ///
		(rcap ci_lower ci_upper e, ///
		lcolor(black) msize(0)) ///
		(scatter coef e, ///
		msym(O) mcolor(black) 	///
		), ///
		legend(off) ///
		xlabel(-5(1)5, nogrid) ///
		ylabel(-20(10)20) ///	
		yline(0, lpattern(dash)) ///
		xline(-1, lpattern(dash)) ///
		xtitle("Event-Time") ///
		ytitle("Treatment Effect" "Mortality Per 100,000", orientation(horizontal) size(small)) ///
		text(19.4 3 "Estimate e ∈ {0,5}=`postcoef'", size(small)) ///
		text(15.5 3 "Std. Error = `postse'" "Conf. Int = [`postlci',`postuci']", size(small))	///
		xsize(10) ysize(5)



```

:::

```{stata, include = FALSE}
*delete all the extra stata files created 
*erase "../data/reg_data_2013.dta"
*erase "../data/reg_data_change.dta"
*erase "../data/csdid_input.dta"
*erase "../data/county_mortality_data_cl.dta"
*erase "../data/data_fig_4.dta"
*erase "../data/figure8.dta"
*erase "../data/precollapse.dta"
*erase "../data/fig6_combined.gph"
*erase "../data/grp2014.gph"
*erase "../data/grp2015.gph"
*erase "../data/grp2016.gph"
*erase "../data/grp2019.gph"
*erase "../data/fig4_dripw.gph"
*erase "../data/fig4_stdipw.gph"
*erase "../data/fig4_reg.gph"

```

```{r, include = FALSE}
#delete all the extra R files we created 
# unlink(xlsx_path)
# unlink(xlsx_path2)
# unlink(xlsx_path3)
# unlink(xlsx_path4)
# unlink(xlsx_path5)
# unlink(xlsx_path6)
# unlink(xlsx_path7)

#unlink(fig_path1)
#unlink(fig_path2)
#unlink(fig_path3)
#unlink(fig_path4)
#unlink(fig_path5)
#unlink(fig_path6)
#unlink(fig_path7)
#unlink(fig_path8)
#unlink(fig_path9)
```
