---
title: "Visualizing Regression with the FWL Theorem in R"
subtitle: "Hands-on tour of fwlplot: confounding, manual FWL, fixed effects, panel data"
author: "Carlos Mendez"
date: "2026-03-27"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: darkly
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 300
execute:
  warning: false
  message: false
---

## 1. Overview

"What does it actually mean to *control for* a variable?" This is perhaps the most common question in applied regression --- and one of the hardest to answer intuitively. When we say "the effect of coupons on sales, controlling for income," we are describing a relationship that lives in multidimensional space and cannot be directly plotted on a 2D scatter plot. Or can it?

The **Frisch-Waugh-Lovell (FWL) theorem** provides the answer. It says that the coefficient on any variable in a multiple regression equals the slope from a simple bivariate regression --- after first "partialling out" the other variables from both the outcome and the variable of interest. Partialling out means regressing a variable on the controls and keeping only the leftover (residual) variation --- the part that the controls cannot explain. This means we *can* visualize any regression coefficient as a 2D scatter plot, as long as we first remove the influence of the controls from both axes.

The [fwlplot](https://cran.r-project.org/package=fwlplot) R package (Butts & McDermott, 2024) turns this into a one-liner. It uses the same formula syntax as [`fixest::feols()`](https://lrberge.github.io/fixest/reference/feols.html) --- including the `|` operator for fixed effects --- and produces a scatter plot of the residualized data with the regression line overlaid. The result is a visual answer to "what does controlling for X look like?"

This tutorial builds intuition progressively. We start with simulated data where we *know* the true effect, show how confounding creates a misleading picture, and use `fwl_plot()` to reveal the truth. We then extend to real data with high-dimensional fixed effects --- first flights data (controlling for origin and destination airports) and then panel wage data (controlling for unobserved individual ability).

**Learning objectives:**

- State the FWL theorem and explain its geometric intuition
- Use `fwl_plot()` to visualize a bivariate relationship before and after controlling for confounders
- Demonstrate that manual FWL residualization reproduces `feols()` coefficients exactly
- Visualize what fixed effects "do" to data by comparing raw vs. residualized scatter plots
- Apply `fwl_plot()` to real panel data with high-dimensional fixed effects
- Connect FWL to omitted variable bias and Simpson's paradox

### Key concepts at a glance

The post leans on a small vocabulary repeatedly. The rest of the tutorial assumes you can move between these terms quickly. Each concept below has three parts: a definition, an example, and an analogy.

**1. Frisch-Waugh-Lovell theorem** $\hat\beta_1 = \hat\beta_1^{\mathrm{resid}}$.
The coefficient on $X_1$ from the full regression equals the slope from a simple regression of $\tilde Y$ on $\tilde X_1$, where the tildes are residuals after partial-ing out the other controls. Two routes, one number.

**Example.** In the simulated store data, regressing `sales` on `coupons` and `income` jointly gives a coupon coefficient of $+0.212$. Manual FWL --- residualize `coupons` against `income`, residualize `sales` against `income`, then regress one residual on the other --- returns $+0.212288$. Same number, two paths.

**Analogy.** Reading "rainfall above normal for the month" instead of "raw rainfall" --- both axes are scrubbed of the seasonal baseline before you compare them.

**2. Partialling out / residualization.**
Replace a variable with the residuals from its regression on the other controls. The residual is the part the controls cannot explain --- the "leftover" variation that drives the FWL coefficient.

**Example.** `resid(lm(coupons ~ income))` strips out income's influence on coupons. The remaining values measure how many coupons each store had *given its income level*. Pair with the same operation on sales and you get the FWL scatter.

**Analogy.** Height-for-age z-scores in pediatrics. You do not compare a 6-year-old to a 12-year-old; you compare each child to other children of the same age.

**3. Omitted variable bias (OVB)** $\hat\beta^{\mathrm{naive}} = \beta + \gamma \cdot \delta$.
A naive regression that omits a confounder is biased by the product of two regressions: the confounder's effect on the outcome ($\gamma$) and on the treatment ($\delta$).

**Example.** In the store data, $\hat\gamma = 0.300$ (income on sales) and $\hat\delta = -0.494$ (income on coupons), so the predicted bias is $-0.148$. The naive coupon coefficient is $-0.093$, the true effect is $+0.212$. Difference: $-0.305$; OVB prediction: $-0.148$; the remainder is sampling noise at $n = 200$.

**Analogy.** Knowing how strong the wind is and how it leans the kite --- you can predict where the kite will land before you let it fly.

**4. Simpson's paradox.**
The direction of an association can reverse when a confounder is added to the model. Two regressions tell *opposite* stories about the same data.

**Example.** Naive `sales ~ coupons` slope: $-0.093$. Add income: slope flips to $+0.212$. Same dataset, opposite signs of effect.

**Analogy.** A medication looks worse in aggregate hospital data because sicker patients get it more often --- once you condition on disease severity, the medication helps.

**5. Fixed effects = demeaning.**
Including a fixed effect for a categorical variable is algebraically identical to subtracting the group mean from every variable in the regression. FWL applied to a set of dummies.

**Example.** In the wage panel, `feols(lwage ~ exper | nr)` is equivalent to first subtracting each person's average `lwage` and average `exper`, then running OLS on the demeaned variables. The slope on `exper` jumps from $0.030$ (pooled) to $0.122$ (within-person).

**Analogy.** Grading on a curve, separately for each classroom. Differences in classroom averages disappear; only deviations from your own classmates' average remain.

**6. Within-person variation.**
The variation that remains after demeaning each individual --- a person's deviations from their own time-averaged outcome. Individual FE isolate this within-person signal.

**Example.** Panel B of the wage figure shows a steep slope of $0.122$ where Panel A shows a flat slope of $0.030$. The within-person return to experience is large; the cross-sectional return is small because cross-sectional differences are dominated by ability.

**Analogy.** A runner's pace today vs. their personal best, ignoring whether they are objectively fast or slow. Personal-best deviations carry the training signal.

## 2. The Modeling Pipeline

```{mermaid}
graph LR
    A["Simulated<br/>Data<br/>(Section 3)"] --> B["fwl_plot()<br/>Naive vs. FWL<br/>(Section 4)"]
    B --> C["Manual FWL<br/>Verification<br/>(Section 5)"]
    C --> D["Fixed Effects<br/>Flights Data<br/>(Section 6)"]
    D --> E["Panel Data<br/>Wages<br/>(Section 7)"]
    E --> F["ggplot2<br/>& Recipe<br/>(Section 8)"]

    style A fill:#6a9bcc,stroke:#141413,color:#fff
    style B fill:#d97757,stroke:#141413,color:#fff
    style C fill:#d97757,stroke:#141413,color:#fff
    style D fill:#6a9bcc,stroke:#141413,color:#fff
    style E fill:#6a9bcc,stroke:#141413,color:#fff
    style F fill:#00d4c8,stroke:#141413,color:#fff
```

We start where the answer is known (simulated data), see the result with `fwl_plot()` first, then peek under the hood with manual FWL verification. From there we apply the same one-liner to increasingly complex real-world settings.

## 3. Setup and Data

### 3.1 Install and load packages

We use `pak::pkg_install()` so the reader gets the **exact** versions used when this post was published. Pinning fosters replicability: a reader who renders this notebook on a fresh machine should get the same numbers, even if upstream packages have changed default behaviour or deprecated functions. `set.seed(42)` is set globally so the simulated data are bit-reproducible.

```{r}
#| label: setup-packages

# pak is the only bootstrap dependency. Everything else gets installed
# at a specific version, so the notebook is reproducible on a fresh
# machine. On a warm machine where the right versions are already
# installed, pak::pkg_install short-circuits and returns quickly.
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", repos = "https://cloud.r-project.org")
}

# Install the EXACT versions used when this post was published.
pak::pkg_install(c(
  "fwlplot@0.3.0",      # FWL scatter plots
  "fixest@0.14.1",      # fast fixed effects (feols)
  "ggplot2@4.0.3",      # plotting
  "patchwork@1.3.2",    # multi-panel composition
  "nycflights13@1.0.2", # real-world flights data
  "wooldridge@1.4.4"    # wage panel data
))

library(fwlplot)
library(fixest)
library(ggplot2)
library(patchwork)
library(nycflights13)
library(wooldridge)

# Site color palette + a shared theme for fwl_plot panels.
steel_blue   <- "#6a9bcc"
warm_orange  <- "#d97757"
near_black   <- "#141413"
teal         <- "#00d4c8"
heading_blue <- "#1a3a8a"

theme_fwl <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(color = heading_blue, face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", size = 11)
  )

set.seed(42)
```

The `fwlplot` package provides the `fwl_plot()` function for FWL-residualized scatter plots. It is built on `fixest`, which handles the residualization computation using fast demeaning algorithms. The `patchwork` package lets us combine multiple ggplot2 plots side by side. The `nycflights13` and `wooldridge` packages provide the real datasets we will use later.

### 3.2 Simulated confounding data

To build intuition, we simulate a retail scenario where a store manager wants to know whether distributing coupons increases sales. The catch: **income is a confounder** --- wealthier neighborhoods receive fewer coupons (the store targets promotions at lower-income areas) but have higher baseline sales. This creates a spurious negative correlation between coupons and sales, even though coupons genuinely boost sales.

The causal structure looks like this:

```{mermaid}
graph TD
    Income["Income<br/>(confounder)"]
    Coupons["Coupons<br/>(treatment)"]
    Sales["Sales<br/>(outcome)"]

    Income -->|"-0.5<br/>(fewer coupons<br/>to rich areas)"| Coupons
    Income -->|"+0.3<br/>(rich areas<br/>buy more)"| Sales
    Coupons -->|"+0.2<br/>(true causal<br/>effect)"| Sales

    style Income fill:#d97757,stroke:#141413,color:#fff
    style Coupons fill:#6a9bcc,stroke:#141413,color:#fff
    style Sales fill:#00d4c8,stroke:#141413,color:#fff
```

Income opens a "backdoor path" from coupons to sales: coupons $\leftarrow$ income $\rightarrow$ sales. Unless we block this path by controlling for income, the naive estimate will be biased. The data generating process is:

$$\text{income} \sim N(50, 10)$$

$$\text{coupons} = 60 - 0.5 \times \text{income} + \epsilon_1, \quad \epsilon_1 \sim N(0, 5)$$

$$\text{sales} = 10 + 0.2 \times \text{coupons} + 0.3 \times \text{income} + \epsilon_2, \quad \epsilon_2 \sim N(0, 3)$$

In words, the true causal effect of coupons on sales is **+0.2**: each additional coupon increases sales by 0.2 units. But because income negatively drives coupons ($-0.5$) and positively drives sales ($+0.3$), a naive regression of sales on coupons alone will confound the coupon effect with the income effect, producing a biased estimate.

```{r}
#| label: data-simulate

set.seed(42)
n <- 200

income    <- rnorm(n, mean = 50, sd = 10)
dayofweek <- sample(1:7, n, replace = TRUE)
coupons   <- 60 - 0.5 * income + rnorm(n, 0, 5)
sales     <- 10 + 0.2 * coupons + 0.3 * income + 0.5 * dayofweek + rnorm(n, 0, 3)

store_data <- data.frame(
  sales     = round(sales, 2),
  coupons   = round(coupons, 2),
  income    = round(income, 2),
  dayofweek = dayofweek
)

head(store_data)
```

```{r}
#| label: data-correlations

round(cor(store_data[, c("sales", "coupons", "income")]), 3)
```

The correlation matrix confirms the confounding structure. Coupons and sales have a *negative* raw correlation ($-0.166$), even though the true causal effect is positive ($+0.2$). This is because income is strongly negatively correlated with coupons ($-0.709$) and strongly positively correlated with sales ($0.500$). A naive analysis would conclude that coupons hurt sales --- a classic instance of **Simpson's paradox**, where the direction of an association reverses when a confounding variable is accounted for.

## 4. fwl_plot() in Action: Naive vs. Controlled

### 4.1 The naive scatter

The simplest way to see why confounding is dangerous: plot the raw relationship with `fwl_plot()`. When no controls are specified, `fwl_plot()` produces a standard scatter plot with a regression line.

The slope is **$-0.093$** ($p = 0.019$): coupons appear to *reduce* sales. This is statistically significant but substantively wrong --- the true effect is $+0.2$. The store manager who trusts this analysis would cancel the coupon program, losing real revenue.

### 4.2 Controlling for income: one line of code

Now watch what happens when we add `income` as a control --- just add it to the formula. The slope reverses to **$+0.212$** ($p < 0.001$) --- close to the true value of $+0.2$. The `fwl_plot()` function residualized both coupons and sales on income behind the scenes, then plotted the residuals.

```{r}
#| label: fig-naive-vs-controlled
#| fig-cap: "Naive scatter (left) shows a negative slope; after FWL residualization on income (right), the slope reverses to positive."
#| fig-width: 12
#| fig-height: 5.5

p1_naive <- fwl_plot(sales ~ coupons, data = store_data, ggplot = TRUE) +
  labs(title = "A. Naive: No Controls",
       x = "Coupons", y = "Sales") +
  theme_fwl

p1_fwl <- fwl_plot(sales ~ coupons + income, data = store_data, ggplot = TRUE) +
  labs(title = "B. FWL: Controlling for Income",
       x = "Coupons (residualized)", y = "Sales (residualized)") +
  theme_fwl

p1_naive + p1_fwl +
  plot_annotation(
    title = "What Does 'Controlling for Income' Look Like?",
    subtitle = "fwl_plot() reveals the true positive effect hidden by confounding"
  )
```

The left panel shows the raw relationship: more coupons, lower sales (a downward slope). The right panel shows the *same* data after removing the influence of income from both axes. Once income is partialled out, the true positive effect of coupons emerges clearly. This is what "controlling for income" looks like geometrically --- and `fwl_plot()` produces it in a single line.

### 4.3 The regression table confirms

The `fixest::feols()` function produces the same coefficient, confirmed by `etable()` for side-by-side comparison:

```{r}
#| label: fit-naive-vs-controlled

fe_naive <- feols(sales ~ coupons, data = store_data)
fe_full  <- feols(sales ~ coupons + income, data = store_data)
etable(fe_naive, fe_full, headers = c("Naive", "Controlled"))
```

Adding income as a control flips the coupon coefficient from $-0.093$ to $+0.212$ and increases the R-squared from $0.028$ to $0.321$. The income coefficient ($0.300$) is close to the true value of $0.3$. Every number in this table corresponds to a visual feature of the `fwl_plot()` scatter plots above.

## 5. Under the Hood: Manual FWL Verification

### 5.1 The three-step recipe

The FWL theorem can be stated as a simple recipe. Think of it like measuring height *for your age*: instead of comparing raw heights, you compare how much taller or shorter each person is than the average for their age group. Similarly, FWL compares how much more or fewer coupons a store had *for its income level*, against how much more or fewer sales it had *for its income level*.

The three steps are:

1. **Regress sales on income**, save the residuals (the part of sales that income cannot explain).
2. **Regress coupons on income**, save the residuals (the part of coupons that income cannot explain).
3. **Regress the sales residuals on the coupon residuals** --- the slope is the coupon coefficient.

```{r}
#| label: fwl-manual

# Step 1: Residualize sales on income.
resid_y <- resid(lm(sales ~ income, data = store_data))
# Step 2: Residualize coupons on income.
resid_x <- resid(lm(coupons ~ income, data = store_data))
# Step 3: Regress residuals on residuals.
fwl_manual <- lm(resid_y ~ resid_x)

cat("feols coefficient:     ",
    round(coef(fe_full)["coupons"], 6), "\n")
cat("Manual FWL coefficient:",
    round(coef(fwl_manual)["resid_x"], 6), "\n")
```

The coefficients match to six decimal places. This is not an approximation --- it is an exact algebraic identity. Every time you run a multiple regression, the software is implicitly performing these three steps for each coefficient.

### 5.2 The formal theorem

For those who want the math, the FWL theorem states that in the regression $Y = X_1 \beta_1 + X_2 \beta_2 + \epsilon$, the coefficient $\hat{\beta}_1$ equals:

$$\hat{\beta}_1 = (\tilde{X}_1' \tilde{X}_1)^{-1} \tilde{X}_1' \tilde{Y}, \quad \text{where} \quad \tilde{Y} = M_{X_2} Y, \quad \tilde{X}_1 = M_{X_2} X_1$$

Here $M_{X_2} = I - X_2(X_2'X_2)^{-1}X_2'$ is the "residual-maker" matrix that projects out the effect of $X_2$. In our example, $Y$ is `sales`, $X_1$ is `coupons`, and $X_2$ is `income`. The tilded variables $\tilde{Y}$ and $\tilde{X}_1$ are the residuals from the `resid()` calls above.

### 5.3 Omitted variable bias: predicting the error

The confounding we saw is not mysterious --- the **omitted variable bias (OVB) formula** predicts it exactly. When we omit income from the regression, the bias on the coupon coefficient is:

$$\text{bias} = \hat{\gamma} \times \hat{\delta}$$

In words, the bias equals the effect of the omitted variable on the outcome ($\hat{\gamma}$) multiplied by the relationship between the omitted variable and the treatment ($\hat{\delta}$). Here $\hat{\gamma}$ is the effect of income on sales (in the full model) and $\hat{\delta}$ is the coefficient from regressing coupons on income.

```{r}
#| label: ovb-calc

gamma_hat <- coef(fe_full)["income"]                              # income -> sales
delta_hat <- coef(lm(coupons ~ income, data = store_data))["income"] # income -> coupons
ovb       <- gamma_hat * delta_hat

cat("gamma (income -> sales):  ", round(gamma_hat, 4), "\n")
cat("delta (income -> coupons):", round(delta_hat, 4), "\n")
cat("OVB = gamma * delta:      ", round(ovb, 4), "\n")
cat("Naive (lm) coefficient:   ", round(coef(fe_naive)["coupons"], 4), "\n")
cat("True (lm controlled):     ", round(coef(fe_full)["coupons"], 4), "\n")
cat("True + OVB (predicted):   ", round(coef(fe_full)["coupons"] + ovb, 4), "\n")
```

The OVB formula predicts a bias of about $-0.148$: income's positive effect on sales ($\hat\gamma = 0.300$) times its negative relationship with coupons ($\hat\delta = -0.494$) produces a large negative bias. The predicted naive coefficient (true + bias $\approx 0.06$) is close to the actual naive coefficient ($-0.09$) --- the small discrepancy comes from sampling variation with $n = 200$. The key insight: the bias is *predictable*. If you know the direction of the confounder's effects on both the treatment and the outcome, you know which way the naive estimate is biased.

### 5.4 Adding more controls

The FWL theorem extends naturally to any number of controls. The `fwl_plot()` call handles it automatically.

```{r}
#| label: fit-three-models

fe_full3 <- feols(sales ~ coupons + income + dayofweek, data = store_data)
etable(fe_naive, fe_full, fe_full3,
       headers = c("Naive", "+ Income", "+ Income + Day"))
```

```{r}
#| label: fig-fwl-verification
#| fig-cap: "Three-panel FWL progression: no controls (left), controlling for income (center), controlling for income + day of week (right)."
#| fig-width: 14
#| fig-height: 5

p2a <- fwl_plot(sales ~ coupons, data = store_data, ggplot = TRUE) +
  labs(title = "A. No controls", x = "Coupons", y = "Sales") + theme_fwl

p2b <- fwl_plot(sales ~ coupons + income, data = store_data, ggplot = TRUE) +
  labs(title = "B. + Income",
       x = "Coupons (resid.)", y = "Sales (resid.)") + theme_fwl

p2c <- fwl_plot(sales ~ coupons + income + dayofweek,
                data = store_data, ggplot = TRUE) +
  labs(title = "C. + Income + Day",
       x = "Coupons (resid.)", y = "Sales (resid.)") + theme_fwl

p2a + p2b + p2c +
  plot_annotation(title = "Progressive Controls: How the Scatter Changes")
```

The coupon coefficient progresses from $-0.093$ (naive, wrong sign), to $+0.212$ (controlling for income), to $+0.222$ (adding day of week). The R-squared jumps from $0.028$ to $0.365$ as we add controls. Each `fwl_plot()` panel shows a tighter cloud as more variation is absorbed by the controls --- the residualized scatter becomes more focused on the *coupon-specific* variation in sales.

## 6. Visualizing Fixed Effects

### 6.1 What are fixed effects?

Fixed effects are a special case of the FWL theorem applied to group dummy variables. When we include airport fixed effects in a regression, we are "partialling out" airport-specific means --- in other words, **demeaning**. Demeaning means subtracting each group's average from every observation in that group. The result is that we compare each airport to *itself* rather than comparing different airports to each other.

Think of it like a race handicap. Raw times compare runners who started at different positions. Demeaning each runner's times converts them to "how much faster or slower than their personal average," making the comparison fair. The FWL theorem guarantees that this demeaning procedure produces the same coefficients as including a full set of dummy variables in the regression.

### 6.2 Flights data: progressive fixed effects

The `nycflights13` dataset contains all domestic flights from New York's three airports (EWR, JFK, LGA) in 2013. We ask: what is the relationship between air time and departure delay?

```{r}
#| label: data-flights

data("flights", package = "nycflights13")
flights_clean <- flights[complete.cases(
  flights[, c("dep_delay", "air_time", "origin", "dest")]
), ]
flights_clean <- flights_clean[flights_clean$dep_delay < 120 &
                                  flights_clean$dep_delay > -30, ]

# Remove singleton origin-dest combos for stable FE estimation.
od_counts <- table(paste(flights_clean$origin, flights_clean$dest))
flights_clean <- flights_clean[
  paste(flights_clean$origin, flights_clean$dest) %in%
    names(od_counts[od_counts > 1]),
]

cat("Observations:  ", nrow(flights_clean), "\n")
cat("Origins:       ", length(unique(flights_clean$origin)), "\n")
cat("Destinations:  ", length(unique(flights_clean$dest)), "\n")
```

We sample 5,000 flights for plotting (the regression line uses all data, only the plotted points are sampled to avoid overplotting):

```{r}
#| label: data-flights-sample

set.seed(123)
flights_sample <- as.data.frame(
  flights_clean[sample(nrow(flights_clean), 5000), ]
)
```

Now the power of `fwl_plot()` --- three one-liners that progressively add fixed effects. In `fixest` syntax, the `|` operator separates regular covariates (left) from fixed effects (right), so `dep_delay ~ air_time | origin + dest` means "regress departure delay on air time, with origin and destination fixed effects".

```{r}
#| label: fig-fixed-effects
#| fig-cap: "Progressive FWL plots: no FE (left), origin FE (center), origin + destination FE (right). Each panel adds more fixed effects, residualizing progressively."
#| fig-width: 14
#| fig-height: 5

p3a <- fwl_plot(dep_delay ~ air_time, data = flights_sample,
                ggplot = TRUE) +
  labs(title = "A. No Fixed Effects",
       x = "Air Time (min)", y = "Dep. Delay (min)") + theme_fwl

p3b <- fwl_plot(dep_delay ~ air_time | origin,
                data = flights_sample, ggplot = TRUE) +
  labs(title = "B. Origin FE",
       x = "Air Time (resid.)", y = "Dep. Delay (resid.)") + theme_fwl

p3c <- fwl_plot(dep_delay ~ air_time | origin + dest,
                data = flights_sample, ggplot = TRUE) +
  labs(title = "C. Origin + Dest FE",
       x = "Air Time (resid.)", y = "Dep. Delay (resid.)") + theme_fwl

p3a + p3b + p3c +
  plot_annotation(
    title = "What Do Fixed Effects 'Do' to the Data?",
    subtitle = "Each panel adds more fixed effects, residualizing progressively"
  )
```

The visual transformation is striking. Panel A (no FE) shows a vague cloud with a nearly flat slope. Panel B (origin FE) removes the three origin-airport means, tightening the horizontal spread. Panel C (origin + destination FE) removes the 103 destination means as well, collapsing the air-time variation to *within-route* deviations.

### 6.3 Comparing regression tables

```{r}
#| label: fit-flights

fe_flights_none   <- feols(dep_delay ~ air_time, data = flights_clean)
fe_flights_origin <- feols(dep_delay ~ air_time | origin, data = flights_clean)
fe_flights_both   <- feols(dep_delay ~ air_time | origin + dest, data = flights_clean)

etable(fe_flights_none, fe_flights_origin, fe_flights_both,
       headers = c("No FE", "Origin FE", "Origin + Dest FE"))
```

The air time coefficient changes as we add fixed effects: $-0.003$ (no FE), $-0.006$ (origin FE), $-0.007$ (origin + destination FE, significant at the 10% level only --- the `.` marker indicates $p < 0.10$). The residualized scatter in Panel C answers a sharper question: "For flights on the *same route*, does longer-than-usual air time predict higher-than-usual departure delay?" The answer is weakly negative --- routes with variable air times show slightly less delay when the flight takes longer, possibly because longer air times reflect favorable wind conditions.

## 7. Panel Data: Returns to Experience

### 7.1 The wage panel

The `wagepan` dataset from the Wooldridge textbook contains panel data on 545 individuals observed over 8 years (1980--1987). A classic question in labor economics is: what is the return to experience?

The challenge is **unobserved ability**. Two people with 5 years of experience may earn very different wages because one is more talented, motivated, or well-connected. These personal traits --- which we cannot directly measure --- are the "unobserved ability" that creates omitted variable bias. More talented workers earn higher wages *and* tend to accumulate experience in higher-paying jobs, so the naive correlation between experience and wages confounds ability with genuine experience effects.

```{r}
#| label: data-wagepan

data("wagepan", package = "wooldridge")
cat("Observations:", nrow(wagepan), "\n")
cat("Individuals:", length(unique(wagepan$nr)), "\n")
cat("Years:", length(unique(wagepan$year)), "\n")
```

### 7.2 Pooled OLS vs. individual fixed effects

```{r}
#| label: fit-wage

fe_pool <- feols(lwage ~ educ + exper + expersq,        data = wagepan)
fe_fe   <- feols(lwage ~ exper + expersq | nr,          data = wagepan)
fe_twfe <- feols(lwage ~ exper + expersq | nr + year,   data = wagepan)

etable(fe_pool, fe_fe, fe_twfe,
       headers = c("Pooled OLS", "Individual FE", "Individual + Year FE"))
```

Several things change as we add fixed effects. First, the `educ` coefficient disappears from the individual FE column --- education is time-invariant for most individuals, so it is perfectly collinear with person dummies. Second, the `exper` linear term disappears from the two-way FE column --- because experience increments by exactly one year for everyone, it is perfectly collinear with year dummies. Only `expersq` (which varies non-linearly across individuals) survives.

In the individual FE model, the experience coefficient *increases* from $0.105$ to $0.122$. This means the within-person return to experience is larger than the pooled estimate. The R-squared jumps from $0.148$ to $0.617$, showing that individual fixed effects explain the majority of wage variation --- most of the "action" in wages comes from *who you are*, not *how many years you have worked*.

### 7.3 Visualizing the within-person variation

Again, `fwl_plot()` produces the before/after comparison in two one-liners. We sample 150 individuals for visual clarity (with 545 individuals the plot would be too dense).

```{r}
#| label: fig-panel-data
#| fig-cap: "Raw pooled cross-section (left) vs. individual fixed-effects residualized scatter (right) for log wage vs. experience."
#| fig-width: 12
#| fig-height: 5.5

set.seed(456)
sample_ids  <- sample(unique(wagepan$nr), 150)
wage_sample <- wagepan[wagepan$nr %in% sample_ids, ]

p4a <- fwl_plot(lwage ~ exper, data = wage_sample, ggplot = TRUE) +
  labs(title = "A. Raw: Pooled Cross-Section",
       x = "Experience (years)", y = "Log Wage") + theme_fwl

p4b <- fwl_plot(lwage ~ exper | nr, data = wage_sample, ggplot = TRUE) +
  labs(title = "B. FWL: Individual Fixed Effects",
       x = "Experience (demeaned)", y = "Log Wage (demeaned)") + theme_fwl

p4a + p4b +
  plot_annotation(
    title = "Controlling for Unobserved Ability",
    subtitle = "Individual FE removes person-specific wage levels, isolating within-person growth"
  )
```

The visual difference is dramatic. Panel A plots the raw bivariate relationship with a shallow slope of about $0.03$. The wide fan of points reflects unobserved ability differences: individuals at the same experience level have wildly different wages. Panel B (individual FE) strips away each person's average wage and average experience, leaving only the *within-person* deviations. The slope steepens to $0.122$ --- more than three times larger --- showing that a one-year increase in experience raises wages by about 12.2% *within the same individual*. The tighter cloud in Panel B shows that once we account for who each person is, the experience-wage relationship is much more precisely identified.

## 8. Customization and Quick Reference

### 8.1 ggplot2 integration

The `fwl_plot()` function can return a ggplot2 object by setting `ggplot = TRUE`, allowing full customization with ggplot2 layers and themes. This is useful for publication-quality figures with consistent styling, faceting, or combining multiple plots with `patchwork`.

```{r}
#| label: fig-ggplot-custom
#| fig-cap: "FWL scatter plot with ggplot2 customization showing coupons effect on sales after residualizing on income."
#| fig-width: 8
#| fig-height: 6

p <- fwl_plot(sales ~ coupons + income, data = store_data, ggplot = TRUE)

p +
  labs(title    = "FWL Visualization: Coupons Effect on Sales",
       subtitle = "After residualizing on income (fwlplot + ggplot2)") +
  theme_fwl
```

### 8.2 Quick reference: fwl_plot() recipes

Here are the most common `fwl_plot()` patterns you will use:

```r
# 1. Raw scatter (no controls)
fwl_plot(y ~ x, data = df)

# 2. Control for one or more variables
fwl_plot(y ~ x + control1 + control2, data = df)

# 3. Fixed effects (use | to separate)
fwl_plot(y ~ x | group_fe, data = df)

# 4. Multiple fixed effects
fwl_plot(y ~ x | fe1 + fe2, data = df)

# 5. Return ggplot2 object for customization
fwl_plot(y ~ x + control, data = df, ggplot = TRUE) + theme_minimal()

# 6. Sample points for large datasets (line uses all data)
fwl_plot(y ~ x | fe, data = big_data, n_sample = 5000)
```

### 8.3 Key arguments

| Argument | Purpose | Example |
|----------|---------|---------|
| `formula` | Same as `feols()`: `y ~ x + controls \| FE` | `sales ~ coupons + income` |
| `data` | Input data frame | `store_data` |
| `ggplot` | Return ggplot2 object (default: base R) | `ggplot = TRUE` |
| `n_sample` | Sample N points for large datasets | `n_sample = 5000` |
| `vcov` | Variance-covariance specification | `vcov = "hetero"` |

For large datasets like the flights data (317K+ observations), the `n_sample` argument is essential to avoid overplotting. The regression line is always computed on the full data --- only the *plotted points* are sampled, so the slope is unaffected.

## 9. Discussion

The FWL theorem is not just a mathematical curiosity --- it is the foundation of how modern regression software works. When `fixest::feols()` estimates a model with fixed effects, it does not literally create and invert a matrix with thousands of dummy variables. Instead, it uses the FWL logic to demean the data and run OLS on the residuals. This is why `fixest` can handle millions of observations with hundreds of thousands of fixed effects: the demeaning step is $O(N)$, while creating the full dummy matrix would be $O(N \times K)$.

As a diagnostic tool, FWL scatter plots reveal problems that regression tables hide. If the residualized scatter shows a curved relationship, your linear specification may be wrong. If it shows outliers, they may be driving the coefficient. If the cloud collapses to a near-vertical line (as in Panel C of the flights figure), the within-group variation may be too small to identify the effect reliably.

The FWL theorem also connects to more advanced methods. **Double Machine Learning** (Chernozhukov et al., 2018) generalizes the partialling-out idea by using machine learning models instead of linear regression to residualize the data. The Python FWL tutorial on this site takes that next step. The `fwlplot` package does not do DML, but the visual intuition --- "look at the residualized scatter to see the conditional relationship" --- carries over directly.

One limitation: the FWL theorem applies only to linear regression. For logistic regression, Poisson regression, or other nonlinear models, the partialling-out logic does not hold exactly. The residualized scatter plot for a nonlinear model is at best an approximation of the conditional relationship, not an exact representation.

## 10. Summary and Next Steps

- **Confounding produces misleading regressions:** in our simulated data, the naive coupon coefficient was $-0.093$ (coupons "hurt" sales), while the true causal effect is $+0.2$. After controlling for income via `fwl_plot()`, the estimate was $+0.212$, recovering the true effect.
- **The OVB formula predicts the bias exactly:** the bias was $0.300 \times (-0.494) = -0.148$, correctly predicting the negative direction and approximate magnitude of the confounding.
- **FWL is not an approximation --- it is an exact algebraic identity:** the coefficient from partialling out controls matches `feols()` to six decimal places. Every multiple regression coefficient *can* be visualized as a bivariate scatter plot.
- **Fixed effects are FWL applied to group dummies:** the flights data showed how adding origin and destination FE progressively transformed the scatter. The air-time coefficient changed from $-0.003$ (no FE) to $-0.007$ (origin + destination FE).
- **Panel FE reveal within-person effects:** the wage data showed that controlling for individual ability via FE steepened the bivariate experience slope from $0.03$ (pooled, no controls) to $0.122$ (within-person), more than tripling the estimated return to experience.

For further study, see the companion [Python FWL tutorial](https://carlos-mendez.org/post/python_fwl/) that extends the partialling-out logic to Double Machine Learning, and the [R DID tutorial](https://carlos-mendez.org/post/r_did/) that uses `fixest` for difference-in-differences with staggered treatment adoption.

## 11. Exercises

1. **Omitted variable direction.** Use the OVB formula from Section 5.3 to predict what happens if you also omit `dayofweek` (in addition to income). Run the naive regression `lm(sales ~ coupons)` and compare the bias to $\hat\gamma_{income} \cdot \hat\delta_{income} + \hat\gamma_{day} \cdot \hat\delta_{day}$. Does the extended OVB formula still predict the direction correctly?

2. **Multiple controls.** Use `fwl_plot()` to visualize the coupon effect after controlling for both income and `dayofweek`. Compare this to controlling for income alone. Does the scatter change visually? Does the coefficient change?

3. **Your own data.** Pick a dataset from the `wooldridge` package (e.g., `hprice1`, `wage2`, `crime2`) and use `fwl_plot()` to visualize a regression relationship before and after adding controls. Does the coefficient change substantially? Can you identify what the confounder is doing?

## 12. References

1. [Butts, K. & McDermott, G. (2024). fwlplot: Scatter Plot After Residualizing. CRAN.](https://cran.r-project.org/package=fwlplot)
2. [Frisch, R. & Waugh, F. V. (1933). Partial Time Regressions as Compared with Individual Trends. *Econometrica*, 1(4), 387--401.](https://doi.org/10.2307/1907330)
3. [Lovell, M. C. (1963). Seasonal Adjustment of Economic Time Series and Multiple Regression Analysis. *JASA*, 58(304), 993--1010.](https://doi.org/10.1080/01621459.1963.10480682)
4. [Berge, L. (2018). fixest: Fast Fixed-Effects Estimations in R. CRAN.](https://cran.r-project.org/package=fixest)
5. [Angrist, J. D. & Pischke, J.-S. (2009). *Mostly Harmless Econometrics.* Princeton University Press.](https://press.princeton.edu/books/paperback/9780691120355/mostly-harmless-econometrics)
6. [Chernozhukov, V. et al. (2018). Double/Debiased Machine Learning for Treatment and Structural Parameters. *The Econometrics Journal*, 21(1), C1--C68.](https://doi.org/10.1111/ectj.12097)

## Source files

- Companion script: [`analysis.R`](https://raw.githubusercontent.com/cmg777/starter-academic-v501/master/content/post/r_fwlplot/analysis.R)
- Published post: <https://carlos-mendez.org/post/r_fwlplot/>
- GitHub repo: <https://github.com/cmg777/starter-academic-v501>
